---
title: "Statistical analysis for Drosophila NSC mosaicism manuscript"
author: "Abhilesh Dhawanjewar"
date: "2026-08-28"
#date: "`r format(Sys.time(), '%Y-%m-%d')`"
format:
    html:
        code-fold: true
        embed-resources: true
        page-layout: full
        toc: true
        toc-depth: 3
        toc-title: "Contents"
        theme: cosmo
execute:
    echo: true
    output: false
    warning: false
---

```{css, echo=FALSE}
p code, li code, td code {
    color: #3a3a3a;
}

.table tbody tr > td,
.table tbody tr > th {
    --bs-table-accent-bg: transparent;
    --bs-table-striped-bg: transparent;
    --bs-table-bg: transparent;
}

.table {
    --bs-table-border-color: #adb5bd;
    border-top-color: #adb5bd;
    border-bottom-color: #adb5bd;
}

.table-striped > tbody > tr:nth-of-type(odd) > td,
.table-striped > tbody > tr:nth-of-type(odd) > th {
    background-color: transparent !important;
}
```

```{r}
#| label: setup
#| code-summary: "Load libraries and data"

library(here)
library(tidyverse)
library(conflicted)
library(lme4)
library(lmerTest)
library(DHARMa)
library(emmeans)
library(kableExtra)
# car, nlme, reactable, htmltools are called via :: only (no attach, no conflicts)

conflicts_prefer(
    dplyr::filter, dplyr::select, dplyr::rename, dplyr::lag,
    lmerTest::lmer, .quiet = TRUE
)

# Reproducibility: a fixed global seed plus deterministic Monte-Carlo quantiles
# for emmeans multivariate-t contrasts (otherwise the p-values vary).
GLOBAL_SEED <- 42L
DHARMA_SEED <- 42L
set.seed(GLOBAL_SEED)
emmeans::emm_options(rng.seed = GLOBAL_SEED)

stat_data_dir    <- here("statistical_analysis", "data")
stat_results_dir <- here("results", "statistical_analysis")
if (!dir.exists(stat_data_dir)) {
    stop("Analysis data not found at ", stat_data_dir)
}
dir.create(stat_results_dir, recursive = TRUE, showWarnings = FALSE)

# Read one per-analysis data table (one row per clone / brain / cell).
read_stat <- function(name) {
    readr::read_csv(file.path(stat_data_dir, name), show_col_types = FALSE)
}

# Write one result table to results/statistical_analysis/.
write_result <- function(x, name) {
    df <- as.data.frame(x)
    rn <- rownames(df)
    if (!is.null(rn) && !identical(rn, as.character(seq_len(nrow(df))))) {
        df <- tibble::rownames_to_column(df, "term")
    }
    readr::write_csv(df, file.path(stat_results_dir, name))
    invisible(df)
}

minimal_tbl <- function(df) {
    df %>%
        kbl(align = "c") %>%
        kable_styling(
            bootstrap_options = c("condensed", "bordered"),
            full_width        = FALSE,
            position          = "left"
        )
}

fmt_p <- function(p, digits = 2) {
    if (is.na(p)) return("NA")
    if (p < 0.001) {
        sci   <- formatC(p, format = "e", digits = digits - 1)
        parts <- strsplit(sci, "e")[[1]]
        exp   <- as.integer(parts[2])
        return(paste0(trimws(parts[1]), " \\times 10^{", exp, "}"))
    }
    as.character(signif(p, max(digits, 2)))
}

sig_stars <- function(p, na_as = "") {
    case_when(
        is.na(p)  ~ na_as,
        p < 0.001 ~ "***",
        p < 0.01  ~ "**",
        p < 0.05  ~ "*",
        TRUE      ~ "ns"
    )
}

sig_tbl <- function(df) {
    sig_col <- intersect(c("sig", "sig_label"), names(df))[1]
    if (is.na(sig_col)) return(minimal_tbl(df))

    sig_vals <- df[[sig_col]]
    rows_3 <- which(sig_vals == "***")
    rows_2 <- which(sig_vals == "**")
    rows_1 <- which(sig_vals == "*")

    df[[sig_col]] <- dplyr::case_when(
        sig_vals == "***" ~ kableExtra::cell_spec(sig_vals, bold = TRUE, color = "#1a7b1a"),
        sig_vals == "**"  ~ kableExtra::cell_spec(sig_vals, bold = TRUE, color = "#2d8a2d"),
        sig_vals == "*"   ~ kableExtra::cell_spec(sig_vals, bold = TRUE, color = "#5aaa5a"),
        sig_vals == "ns"  ~ kableExtra::cell_spec(sig_vals, color = "#999999"),
        TRUE ~ sig_vals
    )

    tbl <- df %>%
        kbl(align = "c", escape = FALSE) %>%
        kable_styling(
            bootstrap_options = c("condensed", "bordered"),
            full_width        = FALSE,
            position          = "left"
        )

    if (length(rows_3) > 0) tbl <- tbl %>% row_spec(rows_3, background = "#c8e6c9")
    if (length(rows_2) > 0) tbl <- tbl %>% row_spec(rows_2, background = "#def0de")
    if (length(rows_1) > 0) tbl <- tbl %>% row_spec(rows_1, background = "#f0faf0")

    tbl
}
```

---

## Fig 1B: NSC lineage size

NSC lineage size (progeny per NSC clone) is compared across Control,
Complex I, and Complex V RNAi, one clone per row. Two mosaic systems affect
different fractions of NSCs and are analysed as separate models:

- **Most NSCs affected:** `worniu-GAL4, Tub-FRT-Stop-FRT-GFP`
- **Few NSCs affected:** `Actin-FRT-Stop-FRT-GAL4, UAS-mCD8-GFP`

Brains are the biological replicate and enter each panel's model as a random
intercept:

`lineage_size ~ rnai + (1 | brain_uid)`

Complex I and Complex V RNAi are each compared to the panel-matched Control RNAi,
with a multivariate-$t$ adjustment across the two comparisons within a panel.

```{r}
#| label: fig1b-load-data
#| code-summary: "Load Fig 1B lineage-size data and tabulate clones and brains per group"
#| output: true

fig1b_df <- read_stat("fig_01b_lineage_size.csv") %>%
    filter(!is.na(value)) %>%
    transmute(
        lineage_size = value,
        genotype     = genotype,
        panel = factor(
            if_else(panel_side == "left", "Most NSCs affected", "Few NSCs affected"),
            levels = c("Most NSCs affected", "Few NSCs affected")
        ),
        rnai      = factor(rnai, levels = c("Control", "Complex I", "Complex V")),
        brain_uid = factor(replicate_id)
    )

fig1b_df %>%
    group_by(panel, rnai) %>%
    summarise(
        n_clones = n(),
        n_brains = n_distinct(brain_uid),
        .groups  = "drop"
    ) %>%
    minimal_tbl()
```

### Diagnostics

```{r}
#| label: fig1b-lme-fit
#| code-summary: "Fit panel-specific linear mixed-effects models"

fig1b_most_df <- fig1b_df %>% filter(panel == "Most NSCs affected")
fig1b_few_df  <- fig1b_df %>% filter(panel == "Few NSCs affected")

fig1b_most_model <- lmerTest::lmer(
    lineage_size ~ rnai + (1 | brain_uid),
    data = fig1b_most_df,
    REML = TRUE
)

fig1b_few_model <- lmerTest::lmer(
    lineage_size ~ rnai + (1 | brain_uid),
    data = fig1b_few_df,
    REML = TRUE
)

summary(fig1b_most_model)
summary(fig1b_few_model)
```

::: {.callout-note appearance="simple" collapse="true"}
## Model fit diagnostics with DHARMa

Scaled residuals from 1,000 simulations are tested for uniform distribution
(Kolmogorov–Smirnov), dispersion, and an excess of outliers; all three should be
non-significant for a well-fitting model.

```{r}
#| label: fig1b-dharma
#| code-summary: "Run DHARMa diagnostics"
#| output: true
#| fig-width: 8
#| fig-height: 5

sim_fig1b_most <- DHARMa::simulateResiduals(fig1b_most_model, n = 1000, seed = DHARMA_SEED)
sim_fig1b_few  <- DHARMa::simulateResiduals(fig1b_few_model,  n = 1000, seed = DHARMA_SEED)

plot(sim_fig1b_most)
plot(sim_fig1b_few)

invisible(capture.output(
    fig1b_dharma_tests <- list(
        most_uniformity     = DHARMa::testUniformity(sim_fig1b_most, plot = FALSE),
        most_dispersion     = DHARMa::testDispersion(sim_fig1b_most, plot = FALSE),
        most_outliers = DHARMa::testOutliers(sim_fig1b_most, plot = FALSE),
        few_uniformity      = DHARMa::testUniformity(sim_fig1b_few,  plot = FALSE),
        few_dispersion      = DHARMa::testDispersion(sim_fig1b_few,  plot = FALSE),
        few_outliers  = DHARMa::testOutliers(sim_fig1b_few,  plot = FALSE)
    )
))

fig1b_dharma_tests %>%
    purrr::map_dfr(
        \(t) tibble(statistic = t$statistic, p.value = t$p.value),
        .id = "test"
    ) %>%
    mutate(across(where(is.numeric), \(x) signif(x, 3)), sig = sig_stars(p.value)) %>%
    minimal_tbl()
```

:::

### Global F-test

Type III ANOVA F-test of the `rnai` term, shown separately for each panel.

```{r}
#| label: fig1b-anova
#| code-summary: "Type III ANOVA"
#| output: true

bind_rows(
    anova(fig1b_most_model, type = "III", ddf = "Satterthwaite") %>%
        as.data.frame() %>%
        tibble::rownames_to_column("term") %>%
        mutate(panel = "Most NSCs affected"),
    anova(fig1b_few_model,  type = "III", ddf = "Satterthwaite") %>%
        as.data.frame() %>%
        tibble::rownames_to_column("term") %>%
        mutate(panel = "Few NSCs affected")
) %>%
    rename(df_num = NumDF, df_den = DenDF, F = `F value`, p.value = `Pr(>F)`) %>%
    select(panel, term, everything()) %>%
    mutate(across(where(is.numeric), \(x) signif(x, 3)), sig = sig_stars(p.value)) %>%
    sig_tbl()
```

### Complex I and Complex V RNAi vs. Control RNAi, per panel (multivariate-$t$ adjusted)

Estimates are in progeny per NSC lineage. Each RNAi condition is compared to
the panel-matched Control RNAi group.

```{r}
#| label: fig1b-contrasts
#| code-summary: "Complex I and Complex V vs. Control contrasts, per panel (multivariate-$t$ adjusted)"

fig1b_most_em <- emmeans::emmeans(fig1b_most_model, specs = ~ rnai)
fig1b_few_em  <- emmeans::emmeans(fig1b_few_model,  specs = ~ rnai)

fig1b_most_contrasts <- summary(
    emmeans::contrast(
        fig1b_most_em,
        method = "trt.vs.ctrl",
        ref    = "Control",
        adjust = "mvt"
    ),
    infer = c(TRUE, TRUE)
) %>%
    as_tibble() %>%
    mutate(panel = "Most NSCs affected")

fig1b_few_contrasts <- summary(
    emmeans::contrast(
        fig1b_few_em,
        method = "trt.vs.ctrl",
        ref    = "Control",
        adjust = "mvt"
    ),
    infer = c(TRUE, TRUE)
) %>%
    as_tibble() %>%
    mutate(panel = "Few NSCs affected")

fig1b_contrast_results <- bind_rows(
    fig1b_most_contrasts,
    fig1b_few_contrasts
) %>%
    select(panel, everything()) %>%
    mutate(sig = sig_stars(p.value))
```

```{r}
#| label: fig1b-contrast-table
#| output: true
#| echo: false

fig1b_contrast_results %>%
    mutate(across(where(is.numeric), \(x) signif(x, 3))) %>%
    sig_tbl()
```

---

## Fig 1G: NSC proliferation vs. in situ heteroplasmy

EdU-positive progeny count is regressed on *in situ* heteroplasmy
(% mutant mtDNA, from *D. melanogaster* / *D. yakuba* mtDNA smFISH), one
neuroblast per row, across four brains (B1–B4) with brain included as a fixed effect:

`progeny ~ heteroplasmy + brain`

```{r}
#| label: fig1g-load-data
#| code-summary: "Load Fig 1G heteroplasmy/EdU data and summarise per brain"
#| output: true

# In situ heteroplasmy is the D. melanogaster / (D. melanogaster + D. yakuba)
# mtDNA smFISH fraction; one row per scored neuroblast, four brains (B1-B4).
fig1g_df <- read_stat("fig_01g_heteroplasmy.csv") %>%
    transmute(
        progeny      = value,
        het_pct      = heteroplasmy_pct,
        het_fraction = heteroplasmy_fraction,
        d_mel, d_yak,
        brain = factor(brain_label, levels = c("B1", "B2", "B3", "B4"))
    )

stopifnot(nrow(fig1g_df) == 53L)

fig1g_df %>%
    group_by(brain) %>%
    summarise(
        n_neuroblasts = n(),
        het_range     = paste0(round(min(het_pct)), "-", round(max(het_pct)), "%"),
        mean_progeny  = round(mean(progeny), 2),
        .groups       = "drop"
    ) %>%
    minimal_tbl()
```

### Diagnostics

```{r}
#| label: fig1g-lm
#| code-summary: "Fit linear model with brain as a fixed effect"

fig1g_lm <- lm(progeny ~ het_pct + brain, data = fig1g_df)
summary(fig1g_lm)
```

::: {.callout-note appearance="simple" collapse="true"}
## Model fit diagnostics with DHARMa

Scaled residuals from 1,000 simulations are tested for uniform distribution (Kolmogorov–Smirnov), dispersion, and an excess of outliers; all three should be non-significant for a well-fitting model.

```{r}
#| label: fig1g-dharma
#| code-summary: "Run DHARMa diagnostics"
#| output: true
#| fig-width: 8
#| fig-height: 5

sim_fig1g <- DHARMa::simulateResiduals(fig1g_lm, n = 1000, seed = DHARMA_SEED)
plot(sim_fig1g)

invisible(capture.output(
    fig1g_dharma_tests <- list(
        uniformity     = DHARMa::testUniformity(sim_fig1g, plot = FALSE),
        dispersion     = DHARMa::testDispersion(sim_fig1g, plot = FALSE),
        outliers = DHARMa::testOutliers(sim_fig1g, plot = FALSE)
    )
))

fig1g_dharma_tests %>%
    purrr::map_dfr(
        \(t) tibble(statistic = t$statistic, p.value = t$p.value),
        .id = "test"
    ) %>%
    mutate(across(where(is.numeric), \(x) signif(x, 3)), sig = sig_stars(p.value)) %>%
    minimal_tbl()
```

:::

### Global F-test

Type III ANOVA on `progeny ~ heteroplasmy + brain`; the `heteroplasmy` term tests whether *in situ* heteroplasmy predicts progeny count once brain is held fixed.

```{r}
#| label: fig1g-anova
#| code-summary: "Type III ANOVA"
#| output: true

fig1g_anova <- car::Anova(fig1g_lm, type = "III") %>%
    as.data.frame() %>%
    tibble::rownames_to_column("term") %>%
    rename(df = Df, F = `F value`, p.value = `Pr(>F)`)

fig1g_anova %>%
    mutate(across(where(is.numeric), \(x) signif(x, 3)), sig = sig_stars(p.value)) %>%
    sig_tbl()
```

```{r}
#| label: fig1g-anova-stats
#| echo: false
#| output: false
fig1g_het_row   <- fig1g_anova %>% filter(term == "het_pct")
fig1g_het_f     <- signif(fig1g_het_row$F, 3)
fig1g_het_df    <- fig1g_het_row$df
fig1g_het_dfres <- fig1g_lm$df.residual
fig1g_het_p     <- fmt_p(fig1g_het_row$p.value)
```

Heteroplasmy has no significant effect on progeny count once brain is accounted for ($F(`r fig1g_het_df`, `r fig1g_het_dfres`) = `r fig1g_het_f`$, $p = `r fig1g_het_p`$).

---

## Fig 1P: Mito-XhoI in heteroplasmic and homoplasmic NSC clones

EdU-positive progeny per NSC clone is compared in a 2 × 2 factorial
of mitochondrially targeted transgene (`MitoGFP` control vs. `MitoXho1`) ×
mtDNA background (`Homoplasmic WT` vs. `Heteroplasmic`), one clone per row, with
brain as the biological replicate.

The data are fit with a linear mixed-effects model:

`EdU ~ mito_type * background + (1 | brain_uid)`

Sum-to-zero contrasts are applied to both factors so that each Type III main
effect is marginal to the interaction. `MitoXho1` is then compared to `MitoGFP`
within each mtDNA background, with a multivariate-$t$ adjustment across the two
comparisons.

```{r}
#| label: fig1p-load-data
#| code-summary: "Load Fig 1P heteroplasmic-clone data and tabulate clones and brains per group"
#| output: true

fig1p_df <- read_stat("fig_01p_heteroplasmic_clones.csv") %>%
    filter(!is.na(value)) %>%
    transmute(
        EdU        = value,
        mito_type  = factor(mito_type,        levels = c("MitoGFP", "MitoXho1")),
        background = factor(mtdna_background,  levels = c("Homoplasmic WT", "Heteroplasmic")),
        brain_uid  = factor(replicate_id)
    )

stopifnot(!anyNA(fig1p_df$mito_type), !anyNA(fig1p_df$background))

fig1p_df %>%
    group_by(mito_type, background) %>%
    summarise(
        n_clones = n(),
        n_brains = n_distinct(brain_uid),
        .groups  = "drop"
    ) %>%
    minimal_tbl()
```

### Diagnostics

```{r}
#| label: fig1p-lme
#| code-summary: "Fit clone-level linear mixed-effects model"

contrasts(fig1p_df$mito_type)  <- contr.sum(2)
contrasts(fig1p_df$background) <- contr.sum(2)

fig1p_lme <- lmerTest::lmer(
    EdU ~ mito_type * background + (1 | brain_uid),
    data = fig1p_df,
    REML = TRUE
)
summary(fig1p_lme)
```

::: {.callout-note appearance="simple" collapse="true"}
## Model fit diagnostics with DHARMa

Scaled residuals from 1,000 simulations are tested for uniform distribution (Kolmogorov–Smirnov), dispersion, and an excess of outliers; all three should be non-significant for a well-fitting model.

```{r}
#| label: fig1p-dharma
#| code-summary: "Run DHARMa diagnostics"
#| output: true
#| fig-width: 8
#| fig-height: 5

sim_fig1p <- DHARMa::simulateResiduals(fig1p_lme, n = 1000, seed = DHARMA_SEED)
plot(sim_fig1p)

invisible(capture.output(
    fig1p_dharma_tests <- list(
        uniformity     = DHARMa::testUniformity(sim_fig1p, plot = FALSE),
        dispersion     = DHARMa::testDispersion(sim_fig1p, plot = FALSE),
        outliers = DHARMa::testOutliers(sim_fig1p, plot = FALSE)
    )
))

fig1p_dharma_tests %>%
    purrr::map_dfr(
        \(t) tibble(statistic = t$statistic, p.value = t$p.value),
        .id = "test"
    ) %>%
    mutate(across(where(is.numeric), \(x) signif(x, 3)), sig = sig_stars(p.value)) %>%
    minimal_tbl()
```

The uniformity (KS) test is borderline
($p = `r signif(fig1p_dharma_tests$uniformity$p.value, 2)`$) but not significant,
and the dispersion and outlier tests are not flagged, so the Gaussian
linear mixed-effects model is retained.

:::

### Global F-test

Type III ANOVA (Satterthwaite degrees of freedom) on the two main effects and their interaction.

```{r}
#| label: fig1p-anova
#| code-summary: "Type III ANOVA"
#| output: true

fig1p_anova <- anova(fig1p_lme, type = "III", ddf = "Satterthwaite") %>%
    as.data.frame()

fig1p_anova %>%
    tibble::rownames_to_column("term") %>%
    rename(df_num = NumDF, df_den = DenDF, F = `F value`, p.value = `Pr(>F)`) %>%
    mutate(across(where(is.numeric), \(x) signif(x, 3)), sig = sig_stars(p.value)) %>%
    sig_tbl()
```

```{r}
#| label: fig1p-anova-stats
#| echo: false
#| output: false
fig1p_bg_f    <- signif(fig1p_anova["background", "F value"], 3)
fig1p_bg_dfd  <- signif(fig1p_anova["background", "DenDF"], 3)
fig1p_bg_p    <- fmt_p(fig1p_anova["background", "Pr(>F)"])
fig1p_mito_p  <- fmt_p(fig1p_anova["mito_type", "Pr(>F)"])
fig1p_int_p   <- fmt_p(fig1p_anova["mito_type:background", "Pr(>F)"])
```

None of the three terms reach significance: `background` ($F(1, `r fig1p_bg_dfd`) = `r fig1p_bg_f`$, $p = `r fig1p_bg_p`$), `mito_type` ($p = `r fig1p_mito_p`$), and their interaction ($p = `r fig1p_int_p`$).

### MitoXho1 vs. MitoGFP, within each mtDNA background (multivariate-$t$ adjusted)

The two within-background comparisons are taken from the full
`mito_type * background` estimated-marginal-means grid, with a multivariate-$t$
adjustment applied jointly across both.

```{r}
#| label: fig1p-contrasts
#| code-summary: "MitoXho1 vs. MitoGFP within each mtDNA background (multivariate-$t$ adjusted)"
#| output: true

fig1p_em <- emmeans::emmeans(fig1p_lme, ~ mito_type * background)

fig1p_contrast_list <- list(
    "MitoXho1 - MitoGFP | Homoplasmic WT" = c(-1, 1,  0, 0),
    "MitoXho1 - MitoGFP | Heteroplasmic"  = c( 0, 0, -1, 1)
)

fig1p_contrast_results <- emmeans::contrast(
    fig1p_em,
    method = fig1p_contrast_list,
    adjust = "mvt"
) %>%
    summary(infer = c(TRUE, TRUE)) %>%
    as_tibble() %>%
    mutate(across(where(is.numeric), \(x) signif(x, 3)), sig = sig_stars(p.value))

fig1p_contrast_results %>%
    sig_tbl()
```

```{r}
#| label: fig1p-contrast-stats
#| echo: false
#| output: false
fig1p_get <- function(bg) fig1p_contrast_results %>% filter(str_detect(contrast, fixed(bg)))
fig1p_homo_est <- fig1p_get("Homoplasmic WT")$estimate
fig1p_homo_p   <- fmt_p(fig1p_get("Homoplasmic WT")$p.value)
fig1p_het_est  <- fig1p_get("Heteroplasmic")$estimate
fig1p_het_p    <- fmt_p(fig1p_get("Heteroplasmic")$p.value)
```

MitoXho1 does not significantly change EdU⁺ progeny count relative to MitoGFP in either the Homoplasmic WT background (`r fig1p_homo_est` progeny, $p = `r fig1p_homo_p`$) or the Heteroplasmic background (`r fig1p_het_est` progeny, $p = `r fig1p_het_p`$).

---

## Fig 3E & Supp Fig 7J: UCP1 / UCP2 rescue experiment

NSC mitotic index (% pH3⁺Dpn⁺ / Dpn⁺) is compared in a 3 × 3
factorial of RNAi condition (`Control`, `Complex I`, `Complex V`) × UCP
expression (`GFP`, `UCP1`, `UCP2`), one brain per row (n = 6–13 per cell). The
`UCP1` analysis is shown in Fig 3E and the `UCP2` analysis is shown in
Supp Fig 7J; both are analysed
here together. An interaction linear model is fit to the per-brain values:

`Mitotic_index ~ rnai * ucp`

```{r}
#| label: load-ucp-data
#| code-summary: "Load UCP1 / UCP2 rescue data and tabulate brains per cell"
#| output: true

ucp_df <- read_stat("fig_03e_ucp_rescue.csv") %>%
    filter(!is.na(value)) %>%
    transmute(
        Mitotic_index = value,
        ucp  = factor(rescue_transgene, levels = c("GFP", "UCP1", "UCP2")),
        rnai = factor(rnai,             levels = c("Control", "Complex I", "Complex V"))
    )

# Explicit sum-to-zero coding is required for interpretable Type III main
# effects in this unbalanced factorial design.
contrasts(ucp_df$ucp)  <- contr.sum(nlevels(ucp_df$ucp))
contrasts(ucp_df$rnai) <- contr.sum(nlevels(ucp_df$rnai))

ucp_df %>%
    count(rnai, ucp, name = "n_brains") %>%
    minimal_tbl()
```

### Diagnostics

```{r}
#| label: ucp-lm-raw
#| code-summary: "Fit interaction linear model (RNAi × UCP) on per-brain mitotic index"

ucp_lm_raw <- lm(Mitotic_index ~ rnai * ucp, data = ucp_df)
summary(ucp_lm_raw)
```

::: {.callout-note appearance="simple" collapse="true"}
## Model fit diagnostics with DHARMa

Scaled residuals from 1,000 simulations are tested for uniform distribution (Kolmogorov–Smirnov), dispersion, and an excess of outliers; all three should be non-significant for a well-fitting model.

```{r}
#| label: ucp-dharma-raw
#| code-summary: "Run DHARMa diagnostics"
#| output: true
#| fig-width: 8
#| fig-height: 5

sim_ucp_raw <- DHARMa::simulateResiduals(ucp_lm_raw, n = 1000, seed = DHARMA_SEED)
plot(sim_ucp_raw)

invisible(capture.output(
    ucp_dharma_tests <- list(
        uniformity     = DHARMa::testUniformity(sim_ucp_raw, plot = FALSE),
        dispersion     = DHARMa::testDispersion(sim_ucp_raw, plot = FALSE),
        outliers = DHARMa::testOutliers(sim_ucp_raw, plot = FALSE)
    )
))

ucp_dharma_tests %>%
    purrr::map_dfr(
        \(t) tibble(statistic = t$statistic, p.value = t$p.value),
        .id = "test"
    ) %>%
    mutate(across(where(is.numeric), \(x) signif(x, 3)), sig = sig_stars(p.value)) %>%
    minimal_tbl()
```

:::

### Global F-test

Type III ANOVA on `rnai * ucp`:

- `rnai`: whether RNAi knockdown affects mitotic index overall
- `ucp`: whether UCP expression alone has an effect
- `rnai:ucp`: whether the UCP effect differs across RNAi backgrounds

```{r}
#| label: ucp-anova
#| code-summary: "Type III ANOVA"
#| output: true

ucp_anova_type3 <- car::Anova(ucp_lm_raw, type = "III")

ucp_anova_type3 %>%
    as.data.frame() %>%
    tibble::rownames_to_column("term") %>%
    rename(df = Df, F = `F value`, p.value = `Pr(>F)`) %>%
    mutate(across(where(is.numeric), \(x) signif(x, 3)), sig = sig_stars(p.value)) %>%
    sig_tbl()
```

```{r}
#| label: ucp-anova-stats
#| echo: false
#| output: false
ucp_av       <- as.data.frame(ucp_anova_type3)
ucp_res_df   <- ucp_av["Residuals", "Df"]
ucp_rnai_f   <- signif(ucp_av["rnai",     "F value"], 3)
ucp_rnai_dfn <- ucp_av["rnai",     "Df"]
ucp_rnai_p   <- fmt_p(ucp_av["rnai",     "Pr(>F)"])
ucp_ucp_f    <- signif(ucp_av["ucp",      "F value"], 3)
ucp_ucp_dfn  <- ucp_av["ucp",      "Df"]
ucp_ucp_p    <- fmt_p(ucp_av["ucp",      "Pr(>F)"])
ucp_int_f    <- signif(ucp_av["rnai:ucp", "F value"], 3)
ucp_int_dfn  <- ucp_av["rnai:ucp", "Df"]
ucp_int_p    <- fmt_p(ucp_av["rnai:ucp", "Pr(>F)"])
```

The RNAi main effect was highly significant ($F(`r ucp_rnai_dfn`, `r ucp_res_df`) = `r ucp_rnai_f`$, $p = `r ucp_rnai_p`$), with knockdown status accounting for the primary variation in mitotic index. The UCP main effect was not significant ($F(`r ucp_ucp_dfn`, `r ucp_res_df`) = `r ucp_ucp_f`$, $p = `r ucp_ucp_p`$), but the interaction was significant ($F(`r ucp_int_dfn`, `r ucp_res_df`) = `r ucp_int_f`$, $p = `r ucp_int_p`$), with the effect of UCP expression varying by RNAi background.

### Within-condition pairwise tests

`UCP1` and `UCP2` vs. their RNAi-matched `GFP` control within each RNAi condition (Dunnett's test, multivariate-$t$ adjustment across the two comparisons per condition). Estimates are in percentage-point units.

```{r}
#| label: ucp-contrasts
#| code-summary: "Marginal means and Dunnett's contrasts (UCP vs. GFP, multivariate-$t$ adjusted)"

ucp_emmeans <- emmeans::emmeans(ucp_lm_raw, ~ ucp | rnai)

ucp_contrasts <- emmeans::contrast(
    ucp_emmeans,
    method = "trt.vs.ctrl",
    ref    = "GFP",
    adjust = "mvt"
)

ucp_contrasts_summary <- summary(ucp_contrasts, infer = c(TRUE, TRUE))
```

```{r}
#| label: ucp-contrasts-table
#| output: true
#| echo: false

ucp_contrasts_summary %>%
    as_tibble() %>%
    mutate(across(where(is.numeric), \(x) signif(x, 3)), sig = sig_stars(p.value)) %>%
    sig_tbl()
```

```{r}
#| label: ucp-contrast-stats
#| echo: false
#| output: false
ucp_cs        <- as_tibble(ucp_contrasts_summary)
ucp_ctrl_ci_p <- min(ucp_cs$p.value[ucp_cs$rnai %in% c("Control", "Complex I")])
ucp_cv1       <- ucp_cs %>% filter(rnai == "Complex V", str_detect(as.character(contrast), "UCP1"))
ucp_cv2       <- ucp_cs %>% filter(rnai == "Complex V", str_detect(as.character(contrast), "UCP2"))
ucp_cv1_est   <- signif(ucp_cv1$estimate, 3)
ucp_cv1_p     <- fmt_p(ucp_cv1$p.value)
ucp_cv2_est   <- signif(ucp_cv2$estimate, 3)
ucp_cv2_p     <- fmt_p(ucp_cv2$p.value)
```

Neither UCP had a significant effect in the `Control` or `Complex I` backgrounds (all $p \ge `r fmt_p(ucp_ctrl_ci_p)`$). In the `Complex V` background, both `UCP1` ($+`r ucp_cv1_est`$ pp, $p = `r ucp_cv1_p`$) and `UCP2` ($+`r ucp_cv2_est`$ pp, $p = `r ucp_cv2_p`$) significantly increased mitotic index relative to `GFP`.

### Cross-condition effect comparisons

Following the significant `rnai:ucp` interaction, these difference-in-differences
contrasts test whether the *magnitude* of the UCP rescue effect differs between
the `Complex V` and `Complex I` backgrounds:

- `(UCP1 − GFP | Complex V) − (UCP1 − GFP | Complex I)`
- `(UCP2 − GFP | Complex V) − (UCP2 − GFP | Complex I)`

```{r}
#| label: ucp-direct-interaction-contrasts
#| code-summary: "UCP rescue effect: (multivariate-$t$ adjusted)"

ucp_cell_emmeans <- emmeans::emmeans(ucp_lm_raw, ~ rnai * ucp)

ucp_all_interactions <- emmeans::contrast(
    ucp_cell_emmeans,
    interaction = c("revpairwise", "trt.vs.ctrl"),
    adjust = "none"
)

ucp_direct_interactions <- subset(
    ucp_all_interactions,
    rnai_revpairwise == "Complex V - Complex I"
)

ucp_direct_interactions_summary <- summary(
    ucp_direct_interactions,
    infer = c(TRUE, TRUE),
    adjust = "mvt"
)

ucp_direct_interactions_summary
```

```{r}
#| label: ucp-interaction-table
#| output: true
#| echo: false

ucp_direct_interactions_summary %>%
    as_tibble() %>%
    mutate(across(where(is.numeric), \(x) signif(x, 3)), sig = sig_stars(p.value)) %>%
    sig_tbl()
```

```{r}
#| label: ucp-interaction-stats
#| echo: false
#| output: false
ucp_di     <- as_tibble(ucp_direct_interactions_summary)
ucp_di_u1  <- ucp_di %>% filter(str_detect(as.character(ucp_trt.vs.ctrl), "UCP1"))
ucp_di_u2  <- ucp_di %>% filter(str_detect(as.character(ucp_trt.vs.ctrl), "UCP2"))
ucp_di_u1_est <- signif(ucp_di_u1$estimate, 3)
ucp_di_u1_p   <- fmt_p(ucp_di_u1$p.value)
ucp_di_u2_est <- signif(ucp_di_u2$estimate, 3)
ucp_di_u2_p   <- fmt_p(ucp_di_u2$p.value)
```

A positive estimate indicates a larger rescue effect in `Complex V`. Multivariate-*t* adjustment is applied across the two contrasts. Both contrasts were significant: the `UCP1` rescue effect was `r ucp_di_u1_est` percentage-points larger in `Complex V` than `Complex I` ($p = `r ucp_di_u1_p`$), and the `UCP2` effect was `r ucp_di_u2_est` percentage-points larger ($p = `r ucp_di_u2_p`$).

---

## Fig 3G: LbNox rescue experiment

NSC mitotic index (% pH3⁺Dpn⁺ / Dpn⁺) is compared in a 3 × 2
factorial of LbNox construct (`Control`, `nlsLbNox`, `mitoLbNox`) × RNAi
background (`Control`, `Complex I`), one brain per row (n = 8–14 per cell). An
interaction linear model is fit to the per-brain values:

`mitotic_index ~ rnai * lbnox`

```{r}
#| label: load-lbnox-data
#| code-summary: "Load Fig 3G LbNox-rescue data and tabulate brains per cell"
#| output: true

# Fig 3G analysing the Control and Complex I backgrounds
lbnox_df <- read_stat("fig_03g_lbnox_rescue.csv") %>%
    filter(!is.na(value), rnai %in% c("Control", "Complex I")) %>%
    transmute(
        mitotic_index = value,
        lbnox = factor(
            if_else(rescue_transgene == "None", "Control", rescue_transgene),
            levels = c("Control", "nlsLbNox", "mitoLbNox")
        ),
        rnai = factor(rnai, levels = c("Control", "Complex I"))
    )

# Sum-to-zero coding for interpretable Type III main effects in this
# unbalanced factorial design.
contrasts(lbnox_df$lbnox) <- contr.sum(nlevels(lbnox_df$lbnox))
contrasts(lbnox_df$rnai)  <- contr.sum(nlevels(lbnox_df$rnai))

lbnox_df %>%
    count(rnai, lbnox, name = "n_brains") %>%
    minimal_tbl()
```

### Diagnostics

```{r}
#| label: lbnox-lm-raw
#| code-summary: "Fit interaction linear model (RNAi × LbNox) on per-brain mitotic index"

lbnox_lm <- lm(mitotic_index ~ rnai * lbnox, data = lbnox_df)
summary(lbnox_lm)
```

::: {.callout-note appearance="simple" collapse="true"}
## Model fit diagnostics with DHARMa

Scaled residuals from 1,000 simulations are tested for uniform distribution (Kolmogorov–Smirnov), dispersion, and an excess of outliers; all three should be non-significant for a well-fitting model.

```{r}
#| label: lbnox-dharma
#| code-summary: "Run DHARMa diagnostics"
#| output: true
#| fig-width: 8
#| fig-height: 5

sim_lbnox <- DHARMa::simulateResiduals(lbnox_lm, n = 1000, seed = DHARMA_SEED)
plot(sim_lbnox)

invisible(capture.output(
    lbnox_dharma_tests <- list(
        uniformity     = DHARMa::testUniformity(sim_lbnox, plot = FALSE),
        dispersion     = DHARMa::testDispersion(sim_lbnox, plot = FALSE),
        outliers = DHARMa::testOutliers(sim_lbnox, plot = FALSE)
    )
))

lbnox_dharma_tests %>%
    purrr::map_dfr(
        \(t) tibble(statistic = t$statistic, p.value = t$p.value),
        .id = "test"
    ) %>%
    mutate(across(where(is.numeric), \(x) signif(x, 3)), sig = sig_stars(p.value)) %>%
    minimal_tbl()
```

:::

### Global F-test

Type III ANOVA on `rnai * lbnox`:

- `rnai`: whether RNAi knockdown affects mitotic index overall
- `lbnox`: whether LbNox expression alone has an effect
- `rnai:lbnox`: whether the LbNox effect differs across RNAi backgrounds

```{r}
#| label: lbnox-anova
#| code-summary: "Type III ANOVA"
#| output: true

lbnox_anova_type3 <- car::Anova(lbnox_lm, type = "III")

lbnox_anova_type3 %>%
    as.data.frame() %>%
    tibble::rownames_to_column("term") %>%
    rename(df = Df, F = `F value`, p.value = `Pr(>F)`) %>%
    mutate(across(where(is.numeric), \(x) signif(x, 3)), sig = sig_stars(p.value)) %>%
    sig_tbl()
```

```{r}
#| label: lbnox-anova-stats
#| echo: false
#| output: false
lb_av       <- as.data.frame(lbnox_anova_type3)
lb_res_df   <- lb_av["Residuals",  "Df"]
lb_rnai_f   <- signif(lb_av["rnai",       "F value"], 3)
lb_rnai_dfn <- lb_av["rnai",       "Df"]
lb_rnai_p   <- fmt_p(lb_av["rnai",       "Pr(>F)"])
lb_lb_f     <- signif(lb_av["lbnox",      "F value"], 3)
lb_lb_dfn   <- lb_av["lbnox",      "Df"]
lb_lb_p     <- fmt_p(lb_av["lbnox",      "Pr(>F)"])
lb_int_f    <- signif(lb_av["rnai:lbnox", "F value"], 3)
lb_int_dfn  <- lb_av["rnai:lbnox", "Df"]
lb_int_p    <- fmt_p(lb_av["rnai:lbnox", "Pr(>F)"])
```

The `rnai` main effect was significant ($F(`r lb_rnai_dfn`, `r lb_res_df`) = `r lb_rnai_f`$, $p = `r lb_rnai_p`$), but the `lbnox` main effect was not ($F(`r lb_lb_dfn`, `r lb_res_df`) = `r lb_lb_f`$, $p = `r lb_lb_p`$). The significant interaction ($F(`r lb_int_dfn`, `r lb_res_df`) = `r lb_int_f`$, $p = `r lb_int_p`$) indicates that the effect of LbNox on mitotic index depends on the RNAi background.

### LbNox rescue within each RNAi background (Dunnett-style, multivariate-$t$ adjusted)

`nlsLbNox` and `mitoLbNox` vs. their RNAi-matched `Control` within each RNAi condition (Dunnett's test, multivariate-$t$ adjustment across the two comparisons per condition). Estimates are in percentage-point units.

```{r}
#| label: lbnox-contrasts
#| code-summary: "Marginal means and Dunnett's contrasts (LbNox vs. Control, multivariate-$t$ adjusted)"

lbnox_emmeans <- emmeans::emmeans(lbnox_lm, ~ lbnox | rnai)

lbnox_contrasts <- emmeans::contrast(
    lbnox_emmeans,
    method = "trt.vs.ctrl",
    ref    = "Control",
    adjust = "mvt"
)

lbnox_contrasts_summary <- summary(lbnox_contrasts, infer = c(TRUE, TRUE))
```

```{r}
#| label: lbnox-contrasts-table
#| output: true
#| echo: false

lbnox_contrasts_summary %>%
    as_tibble() %>%
    mutate(across(where(is.numeric), \(x) signif(x, 3)), sig = sig_stars(p.value)) %>%
    sig_tbl()
```

```{r}
#| label: lbnox-contrast-stats
#| echo: false
#| output: false
lb_cs     <- as_tibble(lbnox_contrasts_summary)
lb_get    <- function(rnai_lvl, construct) {
    lb_cs %>% filter(as.character(rnai) == rnai_lvl,
                     str_detect(as.character(contrast), construct))
}
lb_ctrl_nls_est  <- signif(lb_get("Control",   "nlsLbNox")$estimate, 3)
lb_ctrl_nls_p    <- fmt_p(lb_get("Control",    "nlsLbNox")$p.value)
lb_ctrl_mito_est <- signif(lb_get("Control",   "mitoLbNox")$estimate, 3)
lb_ctrl_mito_p   <- fmt_p(lb_get("Control",    "mitoLbNox")$p.value)
lb_ci_nls_est    <- signif(lb_get("Complex I", "nlsLbNox")$estimate, 3)
lb_ci_nls_p      <- fmt_p(lb_get("Complex I",  "nlsLbNox")$p.value)
lb_ci_mito_est   <- signif(lb_get("Complex I", "mitoLbNox")$estimate, 3)
lb_ci_mito_p     <- fmt_p(lb_get("Complex I",  "mitoLbNox")$p.value)
```

Within the `Control` RNAi background, both nlsLbNox (`r lb_ctrl_nls_est` pp, $p = `r lb_ctrl_nls_p`$) and mitoLbNox (`r lb_ctrl_mito_est` pp, $p = `r lb_ctrl_mito_p`$) slightly reduced mitotic index relative to no-construct controls. Within the `Complex I` RNAi background, both constructs significantly increased mitotic index: nlsLbNox (`r lb_ci_nls_est` pp, $p = `r lb_ci_nls_p`$) and mitoLbNox (`r lb_ci_mito_est` pp, $p = `r lb_ci_mito_p`$).

### Cross-condition effect comparisons

Following the significant `rnai:lbnox` interaction, these difference-in-differences
contrasts test whether the *magnitude* of the LbNox rescue effect differs between
the `Complex I` and `Control` RNAi backgrounds:

- `(nlsLbNox − Control | Complex I) − (nlsLbNox − Control | Control RNAi)`
- `(mitoLbNox − Control | Complex I) − (mitoLbNox − Control | Control RNAi)`

```{r}
#| label: lbnox-interaction-contrasts
#| code-summary: "LbNox rescue effect: difference-in-differences (multivariate-$t$ adjusted)"

lbnox_cell_emmeans <- emmeans::emmeans(lbnox_lm, ~ rnai * lbnox)

lbnox_all_interactions <- emmeans::contrast(
    lbnox_cell_emmeans,
    interaction = c("revpairwise", "trt.vs.ctrl"),
    adjust = "none"
)

lbnox_direct_interactions <- subset(
    lbnox_all_interactions,
    rnai_revpairwise == "Complex I - Control"
)

lbnox_direct_interactions_summary <- summary(
    lbnox_direct_interactions,
    infer = c(TRUE, TRUE),
    adjust = "mvt"
)

lbnox_direct_interactions_summary
```

```{r}
#| label: lbnox-interaction-table
#| output: true
#| echo: false

lbnox_direct_interactions_summary %>%
    as_tibble() %>%
    mutate(across(where(is.numeric), \(x) signif(x, 3)), sig = sig_stars(p.value)) %>%
    sig_tbl()
```

```{r}
#| label: lbnox-interaction-stats
#| echo: false
#| output: false
lb_di      <- as_tibble(lbnox_direct_interactions_summary)
lb_di_nls  <- lb_di %>% filter(str_detect(as.character(lbnox_trt.vs.ctrl), "nlsLbNox"))
lb_di_mito <- lb_di %>% filter(str_detect(as.character(lbnox_trt.vs.ctrl), "mitoLbNox"))
lb_di_nls_est  <- signif(lb_di_nls$estimate, 3)
lb_di_nls_p    <- fmt_p(lb_di_nls$p.value)
lb_di_mito_est <- signif(lb_di_mito$estimate, 3)
lb_di_mito_p   <- fmt_p(lb_di_mito$p.value)
```

A positive estimate indicates a larger rescue effect in `Complex I` relative to `Control` RNAi. Both difference-in-differences contrasts were significant: the nlsLbNox rescue effect was `r lb_di_nls_est` percentage-points larger in the Complex I background ($p = `r lb_di_nls_p`$), and the mitoLbNox effect was `r lb_di_mito_est` percentage-points larger ($p = `r lb_di_mito_p`$).

---

## Fig 4C: Glial Inx2 KD and NSC proliferation

EdU-positive progeny per NSC clone is compared across three
populations produced by a bipartite system that drives clonal Complex I RNAi in
scattered NSCs (LexA/LexOp flip-out) while `repo-GAL4` drives the glial transgene
in all glia of the same brain; one clone per row:

- `Luciferase`: Complex I RNAi NSC clones in control glia (Luciferase RNAi)
- `inx2`: Complex I RNAi NSC clones in Inx2-knockdown glia
- `inx2 (not clones)`: wild-type (non-flipped) NSCs in Inx2-knockdown glia,
  measured from the same brains as the `inx2` clones

`inx2` and `inx2 (not clones)` are measured from the *same two physical brains*;
`Luciferase` is an independent set of three unrelated brains.

Because `inx2` and `inx2 (not clones)` are scored on the same brains, those
two conditions share a random-effect group per brain, while the three
`Luciferase` brains form their own groups:

`EdU ~ condition + (1 | brain_uid)`

The `inx2` vs. `inx2 (not clones)` comparison is therefore within-brain and
well replicated at the clone level, whereas comparisons involving `Luciferase`
are limited by the five independent brains available.

```{r}
#| label: inx2-load-data
#| code-summary: "Load Fig 4C inx2 data and tabulate clones and brains per condition"
#| output: true

# `inx2` and `inx2 (not clones)` are scored on the same physical brains and
# share one `replicate_id`, which is used directly as the random-effect group.
inx2_df <- read_stat("fig_04c_inx2_edu.csv") %>%
    filter(!is.na(value)) %>%
    transmute(
        EdU       = value,
        condition = factor(genotype,
                           levels = c("Luciferase", "inx2", "inx2 (not clones)")),
        brain_uid = replicate_id
    )

inx2_df %>%
    group_by(condition) %>%
    summarise(
        n_clones = n(),
        n_brains = n_distinct(brain_uid),
        .groups  = "drop"
    ) %>%
    minimal_tbl()
```

### Diagnostics

```{r}
#| label: inx2-lme
#| code-summary: "Fit clone-level linear mixed-effects model"

inx2_lme <- lmerTest::lmer(
    EdU ~ condition + (1 | brain_uid),
    data    = inx2_df,
    REML    = TRUE,
    control = lme4::lmerControl(optimizer = "bobyqa")
)

summary(inx2_lme)
```

::: {.callout-note appearance="simple" collapse="true"}
## Model fit diagnostics with DHARMa

Scaled residuals from 1,000 simulations are tested for uniform distribution (Kolmogorov–Smirnov), dispersion, and an excess of outliers; all three should be non-significant for a well-fitting model.

```{r}
#| label: inx2-dharma
#| code-summary: "Run DHARMa diagnostics"
#| output: true
#| fig-width: 8
#| fig-height: 5

sim_inx2 <- DHARMa::simulateResiduals(inx2_lme, n = 1000, seed = DHARMA_SEED)
plot(sim_inx2)

invisible(capture.output(
    inx2_dharma_tests <- list(
        uniformity     = DHARMa::testUniformity(sim_inx2, plot = FALSE),
        dispersion     = DHARMa::testDispersion(sim_inx2, plot = FALSE),
        outliers = DHARMa::testOutliers(sim_inx2, plot = FALSE)
    )
))

inx2_dharma_tests %>%
    purrr::map_dfr(
        \(t) tibble(statistic = t$statistic, p.value = t$p.value),
        .id = "test"
    ) %>%
    mutate(across(where(is.numeric), \(x) signif(x, 3)), sig = sig_stars(p.value)) %>%
    minimal_tbl()
```

:::

```{r}
#| label: inx2-random-effects
#| code-summary: "Variance-components table (between-brain vs. residual)"
#| output: true

vc_inx2 <- as.data.frame(lme4::VarCorr(inx2_lme))
vc_inx2$pct <- round(100 * vc_inx2$vcov / sum(vc_inx2$vcov), 1)

vc_inx2[, c("grp", "vcov", "pct")] %>%
    minimal_tbl()
```

### Global F-test

Type III ANOVA (Satterthwaite degrees of freedom) on the `condition` term, testing whether the three populations differ in EdU incorporation.

```{r}
#| label: inx2-anova
#| code-summary: "Type III ANOVA"
#| output: true

anova(inx2_lme, type = "III", ddf = "Satterthwaite") %>%
    as.data.frame() %>%
    tibble::rownames_to_column("term") %>%
    rename(df_num = NumDF, df_den = DenDF, F = `F value`, p.value = `Pr(>F)`) %>%
    mutate(across(where(is.numeric), \(x) signif(x, 3)), sig = sig_stars(p.value)) %>%
    sig_tbl()
```

### Inx2 KD effect on CI-deficient and WT NSCs (multivariate-$t$ adjusted)

Estimates are in EdU percentage points.

```{r}
#| label: inx2-contrasts
#| code-summary: "Marginal means and the three inx2 contrasts (multivariate-$t$ adjusted)"
#| output: true

inx2_em <- emmeans::emmeans(inx2_lme, specs = ~ condition)

inx2_contrast_list <- list(
    "inx2 - Luciferase"              = c(-1,  1,  0),
    "inx2 (not clones) - inx2"       = c( 0, -1,  1),
    "inx2 (not clones) - Luciferase" = c(-1,  0,  1)
)

inx2_contrast_results <- emmeans::contrast(inx2_em, method = inx2_contrast_list, adjust = "mvt") %>%
    summary(infer = c(TRUE, TRUE)) %>%
    as_tibble() %>%
    mutate(across(where(is.numeric), \(x) signif(x, 3)),
           sig = sig_stars(p.value))

inx2_contrast_results %>%
    sig_tbl()
```

The `inx2 (not clones) - inx2` contrast is a within-brain comparison (both
measured from the same 2 physical brains) and is well powered (df ≈ 35);
contrasts against `Luciferase` are limited by the 5 independent physical
brains available (df ≈ 4) and do not reach significance. WT-neighbor NSCs
(`inx2 (not clones)`) are statistically indistinguishable from the unrelated
WT baseline (`Luciferase`), while the clonal CI-RNAi NSCs (`inx2`)
proliferate significantly less than their WT neighbors in the same brain.

---

## Fig 4D & 4L: Glial RNAi screen (pooled analysis, joint BH-FDR)

EdU-positive progeny per NSC clone is compared across a pooled
set of glial RNAi knockdowns (the transmembrane-transporter and
metabolic-enzyme screens, Fig 4D and Fig 4L, plus the cortex-glia and
glial-OxPhos screen conditions). Each knockdown is compared against the shared
reference controls `mCherry x Luciferase` (wild-type proliferation) and
`ND42 x Luciferase` (Complex I dysfunction baseline), one clone per row.

Brains are the biological replicate and the model is:

`EdU ~ condition + (1 | brain_uid)`

Each knockdown is compared to `ND42 x Luciferase`, with a Benjamini-Hochberg
FDR correction applied jointly across all knockdown contrasts.

```{r}
#| label: combined-load-data
#| code-summary: "Load the pooled glial RNAi screen data"

CTRL_COMB <- "mCherry x Luciferase"
REF_COMB  <- "ND42 x Luciferase"

combined_raw <- read_stat("edu_glial_screen.csv") %>%
    filter(in_combined_screen, !is.na(value)) %>%
    transmute(
        EdU         = value,
        genotype    = genotype,
        date        = collection_date,
        brain_uid   = replicate_id,
        figure      = manuscript_panel,
        screen_type
    )
```

### Data Summary

```{r}
#| label: combined-data-summary
#| code-summary: "Summarise the screen: clones, conditions, knockdowns, brains, dates"
#| output: true

combined_raw %>%
    summarise(
        n_clone_rows = n(),
        n_conditions = n_distinct(genotype),
        n_knockdowns = n_distinct(genotype[screen_type != "control"]),
        n_brains = n_distinct(brain_uid),
        n_dates = n_distinct(date)
    ) %>%
    minimal_tbl()
```

### Statistical analysis

```{r}
#| label: combined-condition-factor
#| code-summary: "List the screen knockdowns and build the condition factor"
#| output: false

screen_kds <- combined_raw %>%
    filter(screen_type != "control") %>%
    pull(genotype) %>%
    unique() %>%
    sort()
n_screen_kds <- length(screen_kds)

combined_model_data <- combined_raw %>%
    mutate(
        condition = factor(
            genotype,
            levels = c(CTRL_COMB, REF_COMB, screen_kds)
        )
    )
```

::: {.callout-note appearance="simple" collapse="true"}
## Replication counts per knockdown

```{r}
#| label: combined-replication-counts
#| output: true
#| echo: false

combined_model_data %>%
    filter(screen_type != "control") %>%
    group_by(figure, screen_type, condition) %>%
    summarise(
        n_clone_rows = n(),
        n_brains     = n_distinct(brain_uid),
        n_dates      = n_distinct(date),
        .groups      = "drop"
    ) %>%
    arrange(figure, screen_type, condition) %>%
    minimal_tbl()
```

:::

### Diagnostics

```{r}
#| label: combined-screen-model
#| code-summary: "Fit combined-screen no-date model"

combined_screen_model <- lmerTest::lmer(
    EdU ~ condition + (1 | brain_uid),
    data = combined_model_data,
    REML = TRUE,
    control = lme4::lmerControl(
        optimizer = "bobyqa",
        optCtrl = list(maxfun = 200000)
    )
)

summary(combined_screen_model)
```

::: {.callout-note appearance="simple" collapse="true"}
## Model fit diagnostics with DHARMa

Scaled residuals from 1,000 simulations are tested for uniform distribution (Kolmogorov–Smirnov), dispersion, and an excess of outliers; all three should be non-significant for a well-fitting model.

```{r}
#| label: combined-dharma
#| code-summary: "Run DHARMa diagnostics"
#| output: true
#| fig-width: 8
#| fig-height: 5

sim_combined <- DHARMa::simulateResiduals(
    combined_screen_model,
    n = 1000,
    seed = DHARMA_SEED
)
plot(sim_combined)

invisible(capture.output(
    combined_dharma_tests <- list(
        uniformity     = DHARMa::testUniformity(sim_combined, plot = FALSE),
        dispersion     = DHARMa::testDispersion(sim_combined, plot = FALSE),
        outliers = DHARMa::testOutliers(sim_combined, plot = FALSE)
    )
))

combined_dharma_tests %>%
    purrr::map_dfr(
        \(t) tibble(statistic = t$statistic, p.value = t$p.value),
        .id = "test"
    ) %>%
    mutate(across(where(is.numeric), \(x) signif(x, 3)), sig = sig_stars(p.value)) %>%
    minimal_tbl()
```

:::

```{r}
#| label: combined-random-effects
#| code-summary: "Variance-components table (between-brain vs. residual)"
#| output: true

vc_combined <- as.data.frame(lme4::VarCorr(combined_screen_model))
vc_combined$pct <- round(100 * vc_combined$vcov / sum(vc_combined$vcov), 1)

vc_combined[, c("grp", "vcov", "pct")] %>%
    minimal_tbl()
```

### Global F-test

Type III ANOVA F-test; the `condition` term tests whether any knockdown differs from the reference across the screen.

```{r}
#| label: combined-anova
#| code-summary: "Type III ANOVA"
#| output: true

combined_anova_res <- anova(combined_screen_model, type = "III", ddf = "Satterthwaite") %>%
    as.data.frame() %>%
    tibble::rownames_to_column("term") %>%
    rename(df_num = NumDF, df_den = DenDF, F = `F value`, p.value = `Pr(>F)`)

combined_anova_res %>%
    mutate(across(where(is.numeric), \(x) signif(x, 3)), sig = sig_stars(p.value)) %>%
    sig_tbl()
```

```{r}
#| label: combined-anova-stats
#| echo: false
#| output: false
comb_f    <- signif(combined_anova_res$F[1], 3)
comb_dfn  <- combined_anova_res$df_num[1]
comb_dfd  <- signif(combined_anova_res$df_den[1], 3)
comb_fp   <- fmt_p(combined_anova_res$p.value[1])
```

The `condition` term was highly significant ($F(`r comb_dfn`, `r comb_dfd`) = `r comb_f`$, $p = `r comb_fp`$); at least one knockdown condition differs from the CI dysfunction baseline.

### Each knockdown vs. ND42 x Luciferase (joint BH-FDR)

Estimates are in EdU percentage points. The phenotype contrast
(`ND42 x Luciferase` vs. `mCherry x Luciferase`) is a single comparison and is
not corrected. Each knockdown is compared to `ND42 x Luciferase`, with a
Benjamini-Hochberg FDR correction applied jointly across all `r n_screen_kds`
knockdown contrasts.

```{r}
#| label: combined-emmeans
#| code-summary: "Estimated marginal means per condition"

combined_emmeans <- emmeans::emmeans(combined_screen_model, specs = ~ condition)
summary(combined_emmeans)
```

```{r}
#| label: combined-phenotype-contrast
#| code-summary: "CI dysfunction vs. wild-type phenotype contrast (uncorrected)"
#| output: true

phenotype_contrast_comb <- emmeans::contrast(
    combined_emmeans,
    method = list(
        "ND42 x Luciferase - mCherry x Luciferase" =
            c(-1, 1, rep(0, length(screen_kds)))
    ),
    adjust = "none"
)

phenotype_contrast_summary <- summary(phenotype_contrast_comb, infer = c(TRUE, TRUE)) %>%
    as_tibble()

phenotype_contrast_summary %>%
    mutate(across(where(is.numeric), \(x) signif(x, 3)), sig = sig_stars(p.value)) %>%
    sig_tbl()
```

```{r}
#| label: combined-phenotype-stats
#| echo: false
#| output: false
pheno_est <- signif(phenotype_contrast_summary$estimate[1], 3)
pheno_t   <- signif(phenotype_contrast_summary$t.ratio[1], 3)
pheno_df  <- signif(phenotype_contrast_summary$df[1], 3)
pheno_p   <- fmt_p(phenotype_contrast_summary$p.value[1])
```

The CI dysfunction baseline (`ND42 x Luciferase`) showed a `r abs(pheno_est)` EdU pp reduction relative to wild-type ($t(`r pheno_df`) = `r pheno_t`$, $p = `r pheno_p`$).

```{r}
#| label: combined-screen-contrasts
#| code-summary: "Each knockdown vs. ND42 x Luciferase; joint Benjamini-Hochberg FDR across knockdowns"
#| output: true

combined_em_screen <- combined_emmeans[
    match(c(REF_COMB, screen_kds), combined_emmeans@grid$condition)
]

combined_screen_contrasts <- summary(
    emmeans::contrast(
        combined_em_screen,
        method = "trt.vs.ctrl",
        ref = 1,
        adjust = "none"
    ),
    infer = c(TRUE, TRUE)
) %>%
    as_tibble() %>%
    mutate(
        gene = as.character(contrast) %>%
            str_remove(fixed(paste0(" - ", REF_COMB))) %>%
            str_remove(fixed(paste0(" / ", REF_COMB))) %>%
            str_replace("^\\((.+)\\)$", "\\1"),
        p.fdr = p.adjust(p.value, method = "BH"),
        sig_label = sig_stars(p.fdr, na_as = NA_character_)
    )

genotype_figure <- combined_raw %>%
    filter(screen_type != "control") %>%
    distinct(genotype, figure, screen_type)

combined_results <- combined_screen_contrasts %>%
    left_join(genotype_figure, by = c("gene" = "genotype")) %>%
    select(
        figure,
        screen_type,
        gene,
        estimate,
        SE,
        df,
        lower.CL,
        upper.CL,
        t.ratio,
        p.value,
        p.fdr,
        sig_label
    ) %>%
    arrange(figure, p.fdr)

combined_results_display <- combined_results %>%
    mutate(across(where(is.numeric), \(x) signif(x, 3)))

reactable::reactable(
    combined_results_display,
    elementId     = "combined_screen_table",
    filterable    = TRUE,
    pagination    = FALSE,
    bordered      = TRUE,
    compact       = TRUE,
    highlight     = TRUE,
    defaultSorted = "p.fdr",
    theme = reactable::reactableTheme(
        borderColor    = "#dee2e6",
        stripedColor   = "#f8f9fa",
        highlightColor = "#f0f0f0",
        cellPadding    = "6px 8px",
        style          = list(fontSize = "0.9rem", fontFamily = "inherit")
    ),
    rowStyle = function(index) {
        sig   <- combined_results_display$sig_label[index]
        color <- switch(sig, "***" = "#c8e6c9", "**" = "#def0de", "*" = "#f0faf0", NA)
        if (!is.na(color)) list(backgroundColor = color) else list()
    },
    defaultColDef = reactable::colDef(
        header = function(value, name) {
            htmltools::tagList(
                value,
                htmltools::tags$span(
                    "⇅",
                    title = "Click to sort",
                    style = "margin-left: 4px; color: #999; font-size: 0.75em; font-weight: normal;"
                )
            )
        }
    ),
    columns = list(
        figure      = reactable::colDef(minWidth = 70),
        screen_type = reactable::colDef(
            name     = "Type",
            minWidth = 90,
            filterInput = function(values, name) {
                htmltools::tags$select(
                    style = "width: 100%; max-width: 100%; box-sizing: border-box;",
                    onchange = sprintf(
                        "Reactable.setFilter('combined_screen_table', '%s', event.target.value || undefined)",
                        name
                    ),
                    htmltools::tags$option(value = "", "All"),
                    lapply(sort(unique(values)), htmltools::tags$option)
                )
            }
        ),
        gene      = reactable::colDef(minWidth = 145),
        estimate  = reactable::colDef(minWidth = 85),
        SE        = reactable::colDef(minWidth = 48),
        df        = reactable::colDef(minWidth = 48),
        lower.CL  = reactable::colDef(name = "Lower", minWidth = 65),
        upper.CL  = reactable::colDef(name = "Upper", minWidth = 65),
        t.ratio   = reactable::colDef(minWidth = 54),
        p.value   = reactable::colDef(minWidth = 70),
        p.fdr     = reactable::colDef(minWidth = 70),
        sig_label = reactable::colDef(name = "Sig.",  minWidth = 55)
    )
)
```

```{r}
#| label: combined-sig-count
#| echo: false
#| output: false
n_sig_combined <- sum(combined_results$p.fdr < 0.05, na.rm = TRUE)
```

`r n_sig_combined` of `r n_screen_kds` knockdowns reached BH-FDR < 0.05 relative to the CI dysfunction baseline.

---

## Fig 4K: Eaat1 transporter rescue experiment

EdU-positive progeny per NSC clone is compared across genotypes that
manipulate glial Eaat1 and NSC RNAi independently (two UAS-driver systems), one
clone per row, with brain as the biological replicate. The genotypes form two
panels:

  - **Left panel (NSC control RNAi):** `mCherry x Luciferase` (NSC ctrl, glia ctrl)
    and `mCherry x Eaat1` (NSC ctrl, glia Eaat1 KD).
  - **Right panel (NSC Complex I KD):** `ND42 x Luciferase` (NSC ND42 KD, glia ctrl),
    `ND42 x Eaat1` (NSC ND42 KD, glia Eaat1 KD), and
    `ND42 x Eaat1 RNAi_Eaat2GFP` (NSC ND42 KD, glia Eaat1 KD + UAS-Eaat2 rescue).

Each panel has its own model. The left panel is fit with a clone-level linear
mixed-effects model with brain as a random intercept,
`EdU ~ condition + (1 | brain_uid)`. The right panel is fit with a generalized
least-squares model on the per-brain mean EdU counts with a condition-specific
variance, `gls(mean_EdU ~ condition, weights = varIdent(form = ~ 1 | condition))`,
giving each condition its own between-brain variance.

Two comparisons, specified a priori from the experimental design, are tested:

1. Eaat1 KD specificity (left panel): `mCherry x Eaat1` vs
`mCherry x Luciferase`: Does glial Eaat1 KD alone affect NSC proliferation?
1. CI dysfunction and Eaat2-GFP rescue (right panel): three related
comparisons, tested together with a joint multiplicity adjustment:
    - `ND42 x Eaat1` vs `ND42 x Luciferase`: Does Eaat1 KD worsen CI dysfunction?
    - `ND42 x Eaat1 RNAi_Eaat2GFP` vs `ND42 x Eaat1`: Does Eaat2 rescue the Eaat1 KD effect?
    - `ND42 x Eaat1 RNAi_Eaat2GFP` vs `ND42 x Luciferase`: Is proliferation restored to the CI-only level?

```{r}
#| label: eaat1-load-data
#| code-summary: "Load Fig 4K Eaat1 data and tabulate clones and brains per condition"
#| output: true

LUC_WT_4K     <- "mCherry x Luciferase"
EAAT1_WT_4K   <- "mCherry x Eaat1"
LUC_CI_4K     <- "ND42 x Luciferase"
EAAT1_CI_4K  <- "ND42 x Eaat1"
EAAT1_RESCUE_4K <- "ND42 x Eaat1 RNAi_Eaat2GFP"

eaat1_left_conditions  <- c(LUC_WT_4K, EAAT1_WT_4K)
eaat1_right_conditions <- c(LUC_CI_4K, EAAT1_CI_4K, EAAT1_RESCUE_4K)
eaat1_all_conditions   <- c(eaat1_left_conditions, eaat1_right_conditions)

eaat1_df <- read_stat("edu_glial_screen.csv") %>%
    filter(!is.na(value), genotype %in% eaat1_all_conditions) %>%
    transmute(
        EdU       = value,
        genotype  = genotype,
        date      = collection_date,
        brain_uid = replicate_id,
        condition = factor(genotype, levels = eaat1_all_conditions)
    )

eaat1_left_df <- eaat1_df %>%
    filter(genotype %in% eaat1_left_conditions) %>%
    mutate(condition = factor(genotype, levels = eaat1_left_conditions))

eaat1_right_df <- eaat1_df %>%
    filter(genotype %in% eaat1_right_conditions) %>%
    mutate(condition = factor(genotype, levels = eaat1_right_conditions))

eaat1_df %>%
    group_by(condition) %>%
    summarise(
        n_clones      = n(),
        n_brains      = n_distinct(brain_uid),
        n_dates       = n_distinct(date),
        .groups       = "drop"
    ) %>%
    minimal_tbl()
```

### Diagnostics

```{r}
#| label: eaat1-lme-fit
#| code-summary: "Fit left-panel LME and right-panel brain-mean GLS"

eaat1_left_model <- lmerTest::lmer(
    EdU ~ condition + (1 | brain_uid),
    data = eaat1_left_df,
    REML = TRUE
)

eaat1_brain_means <- eaat1_df %>%
    group_by(condition, brain_uid) %>%
    summarise(mean_EdU = mean(EdU), .groups = "drop")

eaat1_right_brain_means <- eaat1_brain_means %>%
    filter(condition %in% eaat1_right_conditions) %>%
    mutate(condition = factor(condition, levels = eaat1_right_conditions))

eaat1_right_gls <- nlme::gls(
    mean_EdU ~ condition,
    weights = nlme::varIdent(form = ~1 | condition),
    data    = eaat1_right_brain_means,
    method  = "REML"
)

summary(eaat1_left_model)
summary(eaat1_right_gls)
```

::: {.callout-note appearance="simple" collapse="true"}
## Right panel: raw between-brain variance heterogeneity (Brown–Forsythe test)

A Brown–Forsythe test (Levene's test centred on the median) on the per-brain
means quantifies the between-condition variance heterogeneity that motivates the
heteroscedastic GLS.

```{r}
#| label: eaat1-right-raw-levene
#| code-summary: "Run Brown-Forsythe test on raw brain means"
#| output: true

eaat1_right_brain_means %>%
    group_by(condition) %>%
    summarise(n_brains = n(), sd = signif(sd(mean_EdU), 3), .groups = "drop") %>%
    minimal_tbl()

eaat1_right_raw_levene <- car::leveneTest(
    mean_EdU ~ condition,
    data   = eaat1_right_brain_means,
    center = "median"
)

tibble(
    test      = "Brown-Forsythe (raw brain means)",
    statistic = signif(eaat1_right_raw_levene$`F value`[1], 3),
    p.value   = signif(eaat1_right_raw_levene$`Pr(>F)`[1], 3)
) %>%
    mutate(sig = sig_stars(p.value)) %>%
    minimal_tbl()
```

:::

::: {.callout-note appearance="simple" collapse="true"}
## Left panel: model fit diagnostics with DHARMa

Scaled residuals from 1,000 simulations are tested for uniform distribution (Kolmogorov–Smirnov), dispersion, and an excess of outliers; all three should be non-significant for a well-fitting model.

```{r}
#| label: eaat1-dharma
#| code-summary: "Run DHARMa diagnostics"
#| output: true
#| fig-width: 8
#| fig-height: 5

sim_eaat1_left <- DHARMa::simulateResiduals(eaat1_left_model, n = 1000, seed = DHARMA_SEED)
plot(sim_eaat1_left)

invisible(capture.output(
    eaat1_dharma_tests <- list(
        left_uniformity     = DHARMa::testUniformity(sim_eaat1_left, plot = FALSE),
        left_dispersion     = DHARMa::testDispersion(sim_eaat1_left, plot = FALSE),
        left_outliers = DHARMa::testOutliers(sim_eaat1_left, plot = FALSE)
    )
))

eaat1_dharma_tests %>%
    purrr::map_dfr(
        \(t) tibble(statistic = t$statistic, p.value = t$p.value),
        .id = "test"
    ) %>%
    mutate(across(where(is.numeric), \(x) signif(x, 3)), sig = sig_stars(p.value)) %>%
    minimal_tbl()
```

:::

::: {.callout-note appearance="simple" collapse="true"}
## Right panel: normalized-residual diagnostics (DHARMa does not support `gls` objects)

```{r}
#| label: eaat1-right-gls-diagnostics
#| code-summary: "Normalised-residual normality and homogeneity checks for the GLS"
#| output: true

eaat1_right_gls_resid <- resid(eaat1_right_gls, type = "normalized")
eaat1_right_gls_levene <- car::leveneTest(
    eaat1_right_gls_resid ~ eaat1_right_brain_means$condition,
    center = "median"
)

tibble(
    test      = c("Shapiro-Wilk (normality)", "Levene (homogeneity by condition)"),
    statistic = c(
        signif(shapiro.test(eaat1_right_gls_resid)$statistic, 3),
        signif(eaat1_right_gls_levene$`F value`[1], 3)
    ),
    p.value = c(
        signif(shapiro.test(eaat1_right_gls_resid)$p.value, 3),
        signif(eaat1_right_gls_levene$`Pr(>F)`[1], 3)
    )
) %>%
    mutate(sig = sig_stars(p.value)) %>%
    minimal_tbl()
```

:::

### Global F-test

A Type III ANOVA F-test is shown for the left-panel mixed model. The
right-panel GLS has a single fixed-effect term, so its sequential and marginal
F-tests coincide.

```{r}
#| label: eaat1-anova
#| code-summary: "Global F-tests"
#| output: true

eaat1_anova_left  <- anova(eaat1_left_model, type = "III", ddf = "Satterthwaite") %>% as.data.frame()
eaat1_anova_right <- anova(eaat1_right_gls) %>% as.data.frame()
# anova.gls reports no denominator df; for this single-term model it is
# (n brain-level observations) - (n conditions).
e_right_dfd_val <- nrow(eaat1_right_brain_means) - length(eaat1_right_conditions)

bind_rows(
    eaat1_anova_left %>%
        tibble::rownames_to_column("term") %>%
        transmute(model = "left_panel", term, df_num = NumDF, df_den = DenDF,
                  F = `F value`, p.value = `Pr(>F)`),
    eaat1_anova_right %>%
        tibble::rownames_to_column("term") %>%
        filter(term == "condition") %>%
        transmute(model = "right_panel", term, df_num = numDF, df_den = e_right_dfd_val,
                  F = `F-value`, p.value = `p-value`)
) %>%
    mutate(across(where(is.numeric), \(x) signif(x, 3)), sig = sig_stars(p.value)) %>%
    sig_tbl()
```

```{r}
#| label: eaat1-anova-stats
#| echo: false
#| output: false
e_left_f   <- signif(eaat1_anova_left["condition",  "F value"], 3)
e_left_dfn <- eaat1_anova_left["condition",  "NumDF"]
e_left_dfd <- signif(eaat1_anova_left["condition",  "DenDF"], 3)
e_left_p   <- fmt_p(eaat1_anova_left["condition",  "Pr(>F)"])
e_right_f  <- signif(eaat1_anova_right["condition", "F-value"], 3)
e_right_dfn<- eaat1_anova_right["condition", "numDF"]
e_right_dfd<- e_right_dfd_val
e_right_p  <- fmt_p(eaat1_anova_right["condition", "p-value"])
```

The left-panel model showed no significant condition effect ($F(`r e_left_dfn`, `r e_left_dfd`) = `r e_left_f`$, $p = `r e_left_p`$). The right-panel model was significant ($F(`r e_right_dfn`, `r e_right_dfd`) = `r e_right_f`$, $p = `r e_right_p`$).

### Eaat1 KD specificity and Eaat2-GFP rescue efficacy

All estimates are in EdU percentage points.

```{r}
#| label: eaat1-emmeans
#| code-summary: "Estimated marginal means for the left-panel LME and right-panel GLS"

eaat1_left_em  <- emmeans::emmeans(eaat1_left_model, specs = ~ condition)
# `data =` is required for emmeans() on a nlme::gls fit with a varIdent structure.
eaat1_right_em <- emmeans::emmeans(eaat1_right_gls,  specs = ~ condition,
                                   data = eaat1_right_brain_means)

summary(eaat1_left_em)
summary(eaat1_right_em)
```

#### 1. Eaat1 KD alone (left panel)

`mCherry x Eaat1` vs `mCherry x Luciferase`. Single comparison, no correction.

```{r}
#| label: eaat1-contrast-left
#| code-summary: "Eaat1 KD vs. wild-type, left panel (uncorrected)"
#| output: true

eaat1_left_contrast_summary <- emmeans::contrast(
    eaat1_left_em,
    method = list("mCherry x Eaat1 - mCherry x Luciferase" = c(-1, 1)),
    adjust = "none"
) %>%
    summary(infer = c(TRUE, TRUE)) %>%
    as_tibble()

eaat1_left_contrast_summary %>%
    mutate(across(where(is.numeric), \(x) signif(x, 3)), sig = sig_stars(p.value)) %>%
    sig_tbl()
```

```{r}
#| label: eaat1-contrast-left-stats
#| echo: false
#| output: false
el_est <- signif(eaat1_left_contrast_summary$estimate[1], 3)
el_t   <- signif(eaat1_left_contrast_summary$t.ratio[1], 3)
el_df  <- signif(eaat1_left_contrast_summary$df[1], 3)
el_p   <- fmt_p(eaat1_left_contrast_summary$p.value[1])
```

Glial Eaat1 knockdown alone did not significantly alter NSC proliferation relative to wild-type (`r el_est` pp, $t(`r el_df`) = `r el_t`$, $p = `r el_p`$).

#### 2. CI dysfunction aggravation and Eaat2-GFP rescue (right panel)

Three comparisons from the per-brain GLS, jointly multivariate-$t$ adjusted.
The comparison between `ND42 x Eaat1 RNAi_Eaat2GFP` and `ND42 x Eaat1`
tests whether Eaat2-GFP rescues the glial Eaat1-knockdown effect.

```{r}
#| label: eaat1-contrast-right
#| code-summary: "CI-background Eaat1 KD and Eaat2-GFP rescue contrasts (multivariate-$t$ adjusted)"
#| output: true

eaat1_right_contrasts <- list(
    "ND42 x Eaat1 - ND42 x Luciferase"                  = c(-1,  1,  0),
    "ND42 x Eaat1 RNAi_Eaat2GFP - ND42 x Eaat1"         = c( 0, -1,  1),
    "ND42 x Eaat1 RNAi_Eaat2GFP - ND42 x Luciferase"    = c(-1,  0,  1)
)

eaat1_right_contrast_summary <- emmeans::contrast(
    eaat1_right_em,
    method = eaat1_right_contrasts,
    adjust = "mvt"
) %>%
    summary(infer = c(TRUE, TRUE)) %>%
    as_tibble()

eaat1_right_contrast_summary %>%
    mutate(across(where(is.numeric), \(x) signif(x, 3)), sig = sig_stars(p.value)) %>%
    sig_tbl()
```

```{r}
#| label: eaat1-contrast-right-stats
#| echo: false
#| output: false
er <- eaat1_right_contrast_summary
er_bloom_est <- signif(er$estimate[1], 3)
er_bloom_p   <- fmt_p(er$p.value[1])
er_resc_est  <- signif(er$estimate[2], 3)
er_resc_p    <- fmt_p(er$p.value[2])
er_rest_est  <- signif(er$estimate[3], 3)
er_rest_p    <- fmt_p(er$p.value[3])
```

Glial Eaat1 knockdown significantly worsened NSC proliferation in the CI dysfunction background (`r er_bloom_est` pp, $p = `r er_bloom_p`$). The Eaat2-GFP rescue significantly improved proliferation relative to Eaat1 knockdown alone (`r er_resc_est` pp, $p = `r er_resc_p`$), but proliferation remained significantly below the CI control (`r er_rest_est` pp, $p = `r er_rest_p`$) after multivariate-$t$ correction.

---

## Fig 5E: Ldh rescue experiment

EdU-positive progeny per NSC clone is compared across a glial Ldh knockdown
series with human LDHA, human LDHB, and LbNox rescue constructs, one clone per
row, with brain as the biological replicate. The genotypes form two
panels:

- Left panel (Control RNAi): `mCherry x Luciferase` (WT baseline) and
`mCherry x Ldh` (Ldh knockdown alone).
- Right panel (Complex I RNAi): `ND42 x Luciferase` (ND42 KD alone),
`ND42 x Ldh` (ND42 + Ldh double KD, disease baseline), and three rescue
constructs in the Ldh + ND42 background (`ND42 x LdhA;Ldh`, `ND42 x LdhB;Ldh`, and `ND42 x LbNox;Ldh`).

Each panel has its own model. The left panel is fit with a clone-level linear
mixed-effects model with brain as a random intercept,
`EdU ~ condition + (1 | brain_uid)`. The right panel adds a collection-date
random intercept, `EdU ~ condition + (1 | date_separated) + (1 | brain_uid)`.

Four comparisons, specified a priori from the experimental design, are tested:

  1. Ldh KD alone:`mCherry x Ldh` vs `mCherry x Luciferase`: Does Ldh
knockdown alone affect proliferation?
  2. Ldh worsens ND42: `ND42 x Ldh` vs `ND42 x Luciferase`: Does adding Ldh
knockdown to ND42 further reduce proliferation?
  3. Rescue vs disease baseline: Each rescue construct vs `ND42 x Ldh`, with a
multivariate-$t$ adjustment across the three comparisons.
  4. Rescue to control level: Each rescue construct vs `ND42 x Luciferase`,
with a multivariate-$t$ adjustment across the three comparisons: Is
proliferation restored to the ND42-only level?

```{r}
#| label: load-ldh-data
#| code-summary: "Load Fig 5E Ldh-rescue data and build the condition factors"
#| output: true

LUC_WT  <- "mCherry x Luciferase"
LDH_WT  <- "mCherry x Ldh"
LUC_CI  <- "ND42 x Luciferase"
LDH_CI  <- "ND42 x Ldh"
ldh_rescues <- c("ND42 x LdhA;Ldh", "ND42 x LdhB;Ldh", "ND42 x LbNox;Ldh")

ldh_all_conditions <- c(LUC_WT, LDH_WT, LUC_CI, LDH_CI, ldh_rescues)
ldh_control_rnai_conditions <- c(LUC_WT, LDH_WT)
ldh_complex_i_rescue_conditions <- c(LUC_CI, LDH_CI, ldh_rescues)

ldh_df <- read_stat("edu_glial_screen.csv") %>%
    filter(!is.na(value), genotype %in% ldh_all_conditions) %>%
    transmute(
        EdU       = value,
        genotype  = genotype,
        date      = collection_date,
        brain_uid = replicate_id,
        date_separated = if_else(collection_date == "undated",
                                 paste0("undated__", genotype), collection_date),
        condition = factor(genotype, levels = ldh_all_conditions)
    )

ldh_control_rnai_df <- ldh_df %>%
    filter(genotype %in% ldh_control_rnai_conditions) %>%
    mutate(condition = factor(genotype, levels = ldh_control_rnai_conditions))

ldh_complex_i_rescue_df <- ldh_df %>%
    filter(genotype %in% ldh_complex_i_rescue_conditions) %>%
    mutate(condition = factor(genotype, levels = ldh_complex_i_rescue_conditions))

ldh_df %>%
    group_by(condition) %>%
    summarise(
        n_clones = n(),
        n_brains = n_distinct(brain_uid),
        n_dates = n_distinct(date_separated),
        .groups = "drop"
    ) %>%
    minimal_tbl()
```

### Diagnostics

```{r}
#| label: ldh-lme-fit
#| code-summary: "Fit panel-specific linear mixed-effects models"

ldh_control_rnai_model <- lmerTest::lmer(
    EdU ~ condition + (1 | brain_uid),
    data = ldh_control_rnai_df,
    REML = TRUE
)

ldh_complex_i_rescue_model <- lmerTest::lmer(
    EdU ~ condition + (1 | date_separated) + (1 | brain_uid),
    data = ldh_complex_i_rescue_df,
    REML = TRUE
)

summary(ldh_control_rnai_model)
summary(ldh_complex_i_rescue_model)
```

::: {.callout-note appearance="simple" collapse="true"}
## Model fit diagnostics with DHARMa

Scaled residuals from 1,000 simulations are tested for uniform distribution (Kolmogorov–Smirnov), dispersion, and an excess of outliers; all three should be non-significant for a well-fitting model.

```{r}
#| label: ldh-dharma
#| code-summary: "Run DHARMa diagnostics"
#| output: true
#| fig-width: 8
#| fig-height: 5

sim_ldh_control_rnai <- DHARMa::simulateResiduals(
    ldh_control_rnai_model,
    n = 1000,
    seed = DHARMA_SEED
)
sim_ldh_complex_i_rescue <- DHARMa::simulateResiduals(
    ldh_complex_i_rescue_model,
    n = 1000,
    seed = DHARMA_SEED
)

plot(sim_ldh_control_rnai)
plot(sim_ldh_complex_i_rescue)

invisible(capture.output(
    ldh_dharma_tests <- list(
        control_rnai_uniformity =
            DHARMa::testUniformity(sim_ldh_control_rnai, plot = FALSE),
        control_rnai_dispersion =
            DHARMa::testDispersion(sim_ldh_control_rnai, plot = FALSE),
        control_rnai_outliers =
            DHARMa::testOutliers(sim_ldh_control_rnai, plot = FALSE),
        complex_i_rescue_uniformity =
            DHARMa::testUniformity(sim_ldh_complex_i_rescue, plot = FALSE),
        complex_i_rescue_dispersion =
            DHARMa::testDispersion(sim_ldh_complex_i_rescue, plot = FALSE),
        complex_i_rescue_outliers =
            DHARMa::testOutliers(sim_ldh_complex_i_rescue, plot = FALSE)
    )
))

ldh_dharma_tests %>%
    purrr::map_dfr(
        \(t) tibble(statistic = t$statistic, p.value = t$p.value),
        .id = "test"
    ) %>%
    mutate(across(where(is.numeric), \(x) signif(x, 3)), sig = sig_stars(p.value)) %>%
    minimal_tbl()
```

:::

### Global F-test

Type III ANOVA F-tests are shown separately for the
left- and right-panel models.

```{r}
#| label: ldh-anova
#| code-summary: "Type III ANOVA"
#| output: true

ldh_anova_ctrl <- anova(ldh_control_rnai_model,    type = "III", ddf = "Satterthwaite") %>% as.data.frame()
ldh_anova_ci   <- anova(ldh_complex_i_rescue_model, type = "III", ddf = "Satterthwaite") %>% as.data.frame()

bind_rows(
    ldh_anova_ctrl %>% tibble::rownames_to_column("term") %>% mutate(model = "control_rnai"),
    ldh_anova_ci   %>% tibble::rownames_to_column("term") %>% mutate(model = "complex_i_rescue")
) %>%
    rename(df_num = NumDF, df_den = DenDF, F = `F value`, p.value = `Pr(>F)`) %>%
    select(model, term, everything()) %>%
    mutate(across(where(is.numeric), \(x) signif(x, 3)), sig = sig_stars(p.value)) %>%
    sig_tbl()
```

```{r}
#| label: ldh-anova-stats
#| echo: false
#| output: false
ldh_ctrl_f   <- signif(ldh_anova_ctrl["condition", "F value"], 3)
ldh_ctrl_dfn <- ldh_anova_ctrl["condition", "NumDF"]
ldh_ctrl_dfd <- signif(ldh_anova_ctrl["condition", "DenDF"], 3)
ldh_ctrl_p   <- fmt_p(ldh_anova_ctrl["condition", "Pr(>F)"])
ldh_ci_f     <- signif(ldh_anova_ci["condition",   "F value"], 3)
ldh_ci_dfn   <- ldh_anova_ci["condition",   "NumDF"]
ldh_ci_dfd   <- signif(ldh_anova_ci["condition",   "DenDF"], 3)
ldh_ci_p     <- fmt_p(ldh_anova_ci["condition",   "Pr(>F)"])
```

The left-panel model (Control RNAi background) showed no significant condition effect ($F(`r ldh_ctrl_dfn`, `r ldh_ctrl_dfd`) = `r ldh_ctrl_f`$, $p = `r ldh_ctrl_p`$). The right-panel model (CI rescue conditions) was highly significant ($F(`r ldh_ci_dfn`, `r ldh_ci_dfd`) = `r ldh_ci_f`$, $p = `r ldh_ci_p`$).

### Ldh knockdown effect, Complex I aggravation, and rescue efficacy

All estimates are in EdU percentage points.

```{r}
#| label: ldh-emmeans
#| code-summary: "Estimated marginal means for the control-RNAi and CI-rescue panels"

ldh_control_rnai_em <- emmeans::emmeans(
    ldh_control_rnai_model,
    specs = ~ condition
)
ldh_complex_i_rescue_em <- emmeans::emmeans(
    ldh_complex_i_rescue_model,
    specs = ~ condition
)

summary(ldh_control_rnai_em)
summary(ldh_complex_i_rescue_em)
```

#### 1. Ldh knockdown alone (left panel)

`mCherry x Ldh` vs `mCherry x Luciferase`. Single comparison, no correction.

```{r}
#| label: ldh-contrast-ldh-alone
#| code-summary: "Ldh KD vs. wild-type, left panel"
#| output: true

ldh_alone_summary <- emmeans::contrast(
    ldh_control_rnai_em,
    method = list("mCherry x Ldh - mCherry x Luciferase" = c(-1, 1)),
    adjust = "none"
) %>%
    summary(infer = c(TRUE, TRUE)) %>%
    as_tibble()

ldh_alone_summary %>%
    mutate(across(where(is.numeric), \(x) signif(x, 3)), sig = sig_stars(p.value)) %>%
    sig_tbl()
```

```{r}
#| label: ldh-alone-stats
#| echo: false
#| output: false
ldh_alone_est <- signif(ldh_alone_summary$estimate[1], 3)
ldh_alone_t   <- signif(ldh_alone_summary$t.ratio[1], 3)
ldh_alone_df  <- signif(ldh_alone_summary$df[1], 3)
ldh_alone_p   <- fmt_p(ldh_alone_summary$p.value[1])
```

Ldh knockdown alone did not significantly alter NSC proliferation relative to wild-type (`r ldh_alone_est` pp, $t(`r ldh_alone_df`) = `r ldh_alone_t`$, $p = `r ldh_alone_p`$).

#### 2. Ldh KD worsens ND42 phenotype (right panel)

`ND42 x Ldh` vs `ND42 x Luciferase`. Single comparison, no correction.

```{r}
#| label: ldh-contrast-ldh-worsens
#| code-summary: "Ldh KD + CI vs. CI alone, right panel"
#| output: true

ldh_worsens_summary <- emmeans::contrast(
    ldh_complex_i_rescue_em,
    method = list("ND42 x Ldh - ND42 x Luciferase" = c(-1, 1, 0, 0, 0)),
    adjust = "none"
) %>%
    summary(infer = c(TRUE, TRUE)) %>%
    as_tibble()

ldh_worsens_summary %>%
    mutate(across(where(is.numeric), \(x) signif(x, 3)), sig = sig_stars(p.value)) %>%
    sig_tbl()
```

```{r}
#| label: ldh-worsens-stats
#| echo: false
#| output: false
ldh_w_est <- signif(ldh_worsens_summary$estimate[1], 3)
ldh_w_t   <- signif(ldh_worsens_summary$t.ratio[1], 3)
ldh_w_df  <- signif(ldh_worsens_summary$df[1], 3)
ldh_w_p   <- fmt_p(ldh_worsens_summary$p.value[1])
```

Combined Ldh knockdown with CI dysfunction significantly reduced NSC proliferation relative to CI dysfunction alone (`r ldh_w_est` pp, $t(`r ldh_w_df`) = `r ldh_w_t`$, $p = `r ldh_w_p`$).

#### 3. Rescue vs disease baseline

Each rescue construct vs `ND42 x Ldh` (multivariate-$t$ adjusted across 3 comparisons).

```{r}
#| label: ldh-rescue-vs-baseline
#| code-summary: "Rescue vs ND42 x Ldh (multivariate-$t$ adjusted)"

ldh_em_vs_baseline <- ldh_complex_i_rescue_em[
    match(c(LDH_CI, ldh_rescues), ldh_complex_i_rescue_em@grid$condition)
]

ldh_rescue_vs_baseline_summary <- summary(
    emmeans::contrast(ldh_em_vs_baseline, method = "trt.vs.ctrl", ref = 1, adjust = "mvt"),
    infer = c(TRUE, TRUE)
)
```

```{r}
#| label: ldh-rescue-vs-baseline-table
#| output: true
#| echo: false

ldh_rescue_vs_baseline_summary %>%
    as_tibble() %>%
    mutate(
        contrast = str_remove(contrast, fixed(paste0(" - ", LDH_CI))),
        across(where(is.numeric), \(x) signif(x, 3)),
        sig = sig_stars(p.value)
    ) %>%
    sig_tbl()
```

```{r}
#| label: ldh-rescue-vs-baseline-stats
#| echo: false
#| output: false
ldh_vsb       <- as_tibble(ldh_rescue_vs_baseline_summary) %>%
    mutate(gene = str_remove(contrast, fixed(paste0(" - ", LDH_CI))))
ldh_vsb_sig   <- ldh_vsb %>% filter(p.value < 0.05) %>% pull(gene)
ldh_vsb_ns    <- ldh_vsb %>% filter(p.value >= 0.05) %>% pull(gene)
ldh_vsb_sig_str <- if (length(ldh_vsb_sig) == 0) "None" else paste(ldh_vsb_sig, collapse = " and ")
ldh_vsb_ns_str  <- if (length(ldh_vsb_ns)  == 0) "none" else paste(ldh_vsb_ns,  collapse = " and ")
```

`r ldh_vsb_sig_str` significantly increased proliferation relative to the `ND42 x Ldh` baseline; `r ldh_vsb_ns_str` did not.

#### 4. Rescue to control level

Each rescue construct vs `ND42 x Luciferase` (multivariate-$t$ adjusted across 3 comparisons).
A non-significant result indicates the condition is not detectably different from
the ND42-only control, i.e. proliferation is restored.

```{r}
#| label: ldh-rescue-vs-control
#| code-summary: "Rescue vs ND42 x Luciferase (multivariate-$t$ adjusted)"

ldh_em_vs_control <- ldh_complex_i_rescue_em[
    match(c(LUC_CI, ldh_rescues), ldh_complex_i_rescue_em@grid$condition)
]

ldh_rescue_vs_control_summary <- summary(
    emmeans::contrast(ldh_em_vs_control, method = "trt.vs.ctrl", ref = 1, adjust = "mvt"),
    infer = c(TRUE, TRUE)
)
```

```{r}
#| label: ldh-rescue-vs-control-table
#| output: true
#| echo: false

ldh_rescue_vs_control_summary %>%
    as_tibble() %>%
    mutate(
        contrast = str_remove(contrast, fixed(paste0(" - ", LUC_CI))),
        across(where(is.numeric), \(x) signif(x, 3)),
        sig = sig_stars(p.value)
    ) %>%
    sig_tbl()
```

```{r}
#| label: ldh-rescue-vs-control-stats
#| echo: false
#| output: false
ldh_vsc     <- as_tibble(ldh_rescue_vs_control_summary) %>%
    mutate(gene = str_remove(contrast, fixed(paste0(" - ", LUC_CI))))
ldh_vsc_ns  <- ldh_vsc %>% filter(p.value >= 0.05) %>% pull(gene)
ldh_vsc_sig <- ldh_vsc %>% filter(p.value <  0.05) %>% pull(gene)
ldh_vsc_ns_str  <- if (length(ldh_vsc_ns)  == 0) "None" else paste(ldh_vsc_ns,  collapse = " and ")
ldh_vsc_sig_str <- if (length(ldh_vsc_sig) == 0) "none" else paste(ldh_vsc_sig, collapse = " and ")
```

`r ldh_vsc_ns_str` remained indistinguishable from the `ND42 x Luciferase` control (partial restoration); `r ldh_vsc_sig_str` was significantly below the control level.

---

## Fig S1I-J: NSC proliferation rate

NSC proliferation rate (EdU-positive progeny per clone) is
compared across Control, Complex I, and Complex V RNAi, one clone per row. Two
mosaic systems affect different fractions of NSCs and are analysed as separate
models:

- **Most NSCs affected:** `worniu-GAL4, Tub-FRT-Stop-FRT-GFP`
- **Few NSCs affected:** `Actin-FRT-Stop-FRT-GAL4, UAS-mCD8-GFP`

Brains are the biological replicate and enter each panel's model as a random
intercept:

`EdU ~ rnai + (1 | brain_uid)`

Complex I and Complex V RNAi are each compared to the panel-matched Control RNAi,
with a multivariate-$t$ adjustment across the two comparisons within a panel.

```{r}
#| label: load-s1e-data
#| code-summary: "Load Fig S1I-J proliferation-rate data and tabulate clones and brains per group"
#| output: true

s1e_df <- read_stat("fig_s01i_j_edu.csv") %>%
    filter(!is.na(value)) %>%
    transmute(
        EdU   = value,
        panel = factor(
            if_else(panel_side == "left", "Most NSCs affected", "Few NSCs affected"),
            levels = c("Most NSCs affected", "Few NSCs affected")
        ),
        rnai      = factor(rnai, levels = c("Control", "Complex I", "Complex V")),
        brain_uid = factor(replicate_id)
    )

s1e_df %>%
    group_by(panel, rnai) %>%
    summarise(
        n_clones = n(),
        n_brains = n_distinct(brain_uid),
        .groups = "drop"
    ) %>%
    minimal_tbl()
```

### Diagnostics

```{r}
#| label: s1e-lme-fit
#| code-summary: "Fit panel-specific linear mixed-effects models"

s1e_most_nscs_df <- s1e_df %>%
    filter(panel == "Most NSCs affected")

s1e_few_nscs_df <- s1e_df %>%
    filter(panel == "Few NSCs affected")

s1e_most_nscs_model <- lmerTest::lmer(
    EdU ~ rnai + (1 | brain_uid),
    data = s1e_most_nscs_df,
    REML = TRUE
)

s1e_few_nscs_model <- lmerTest::lmer(
    EdU ~ rnai + (1 | brain_uid),
    data = s1e_few_nscs_df,
    REML = TRUE
)

summary(s1e_most_nscs_model)
summary(s1e_few_nscs_model)
```

::: {.callout-note appearance="simple" collapse="true"}
## Model fit diagnostics with DHARMa

Scaled residuals from 1,000 simulations are tested for uniform distribution (Kolmogorov–Smirnov), dispersion, and an excess of outliers; all three should be non-significant for a well-fitting model.

```{r}
#| label: s1e-dharma
#| code-summary: "Run DHARMa diagnostics"
#| output: true
#| fig-width: 8
#| fig-height: 5

sim_s1e_most_nscs <- DHARMa::simulateResiduals(
    s1e_most_nscs_model,
    n = 1000,
    seed = DHARMA_SEED
)
sim_s1e_few_nscs <- DHARMa::simulateResiduals(
    s1e_few_nscs_model,
    n = 1000,
    seed = DHARMA_SEED
)

plot(sim_s1e_most_nscs)
plot(sim_s1e_few_nscs)

invisible(capture.output(
    s1e_dharma_tests <- list(
        most_nscs_uniformity =
            DHARMa::testUniformity(sim_s1e_most_nscs, plot = FALSE),
        most_nscs_dispersion =
            DHARMa::testDispersion(sim_s1e_most_nscs, plot = FALSE),
        most_nscs_outliers =
            DHARMa::testOutliers(sim_s1e_most_nscs, plot = FALSE),
        few_nscs_uniformity =
            DHARMa::testUniformity(sim_s1e_few_nscs, plot = FALSE),
        few_nscs_dispersion =
            DHARMa::testDispersion(sim_s1e_few_nscs, plot = FALSE),
        few_nscs_outliers =
            DHARMa::testOutliers(sim_s1e_few_nscs, plot = FALSE)
    )
))

s1e_dharma_tests %>%
    purrr::map_dfr(
        \(t) tibble(statistic = t$statistic, p.value = t$p.value),
        .id = "test"
    ) %>%
    mutate(across(where(is.numeric), \(x) signif(x, 3)), sig = sig_stars(p.value)) %>%
    minimal_tbl()
```

:::

### Global F-test

Type III ANOVA F-tests are shown separately for the two panel-specific models.

```{r}
#| label: s1e-anova
#| code-summary: "Type III ANOVA"
#| output: true

s1e_anova_most <- anova(s1e_most_nscs_model, type = "III", ddf = "Satterthwaite") %>% as.data.frame()
s1e_anova_few  <- anova(s1e_few_nscs_model,  type = "III", ddf = "Satterthwaite") %>% as.data.frame()

bind_rows(
    s1e_anova_most %>% tibble::rownames_to_column("term") %>% mutate(panel = "Most NSCs affected"),
    s1e_anova_few  %>% tibble::rownames_to_column("term") %>% mutate(panel = "Few NSCs affected")
) %>%
    rename(df_num = NumDF, df_den = DenDF, F = `F value`, p.value = `Pr(>F)`) %>%
    select(panel, term, everything()) %>%
    mutate(across(where(is.numeric), \(x) signif(x, 3)), sig = sig_stars(p.value)) %>%
    sig_tbl()
```

```{r}
#| label: s1e-anova-stats
#| echo: false
#| output: false
s1e_most_f   <- signif(s1e_anova_most["rnai", "F value"], 3)
s1e_most_dfn <- s1e_anova_most["rnai", "NumDF"]
s1e_most_dfd <- signif(s1e_anova_most["rnai", "DenDF"], 3)
s1e_most_p   <- fmt_p(s1e_anova_most["rnai", "Pr(>F)"])
s1e_few_f    <- signif(s1e_anova_few["rnai",  "F value"], 3)
s1e_few_dfn  <- s1e_anova_few["rnai",  "NumDF"]
s1e_few_dfd  <- signif(s1e_anova_few["rnai",  "DenDF"], 3)
s1e_few_p    <- fmt_p(s1e_anova_few["rnai",  "Pr(>F)"])
```

The `rnai` main effect was highly significant in the Most NSCs panel ($F(`r s1e_most_dfn`, `r s1e_most_dfd`) = `r s1e_most_f`$, $p = `r s1e_most_p`$) but not in the Few NSCs panel ($F(`r s1e_few_dfn`, `r s1e_few_dfd`) = `r s1e_few_f`$, $p = `r s1e_few_p`$).

### Complex I and Complex V RNAi vs. Control RNAi, per panel (multivariate-$t$ adjusted)

Estimates are in EdU percentage points. Each RNAi condition is compared to the
panel-matched Control RNAi group.

```{r}
#| label: s1e-contrasts
#| code-summary: "Complex I and Complex V vs. Control contrasts, per panel (multivariate-$t$ adjusted)"

s1e_most_nscs_em <- emmeans::emmeans(s1e_most_nscs_model, specs = ~ rnai)
s1e_few_nscs_em <- emmeans::emmeans(s1e_few_nscs_model, specs = ~ rnai)

s1e_most_nscs_contrasts <- summary(
    emmeans::contrast(
        s1e_most_nscs_em,
        method = "trt.vs.ctrl",
        ref = "Control",
        adjust = "mvt"
    ),
    infer = c(TRUE, TRUE)
) %>%
    as_tibble() %>%
    mutate(panel = "Most NSCs affected")

s1e_few_nscs_contrasts <- summary(
    emmeans::contrast(
        s1e_few_nscs_em,
        method = "trt.vs.ctrl",
        ref = "Control",
        adjust = "mvt"
    ),
    infer = c(TRUE, TRUE)
) %>%
    as_tibble() %>%
    mutate(panel = "Few NSCs affected")

s1e_contrast_results <- bind_rows(
    s1e_most_nscs_contrasts,
    s1e_few_nscs_contrasts
) %>%
    select(panel, everything()) %>%
    mutate(sig = sig_stars(p.value))
```

```{r}
#| label: s1e-contrast-table
#| output: true
#| echo: false

s1e_contrast_results %>%
    mutate(across(where(is.numeric), \(x) signif(x, 3))) %>%
    sig_tbl()
```

```{r}
#| label: s1e-contrast-stats
#| echo: false
#| output: false
s1e_get <- function(panel_name, contrast_name) {
    s1e_contrast_results %>%
        filter(panel == panel_name, str_detect(contrast, contrast_name))
}
s1e_most_ci_est <- signif(s1e_get("Most NSCs affected", "Complex I")$estimate, 3)
s1e_most_ci_p   <- fmt_p(s1e_get("Most NSCs affected", "Complex I")$p.value)
s1e_most_cv_est <- signif(s1e_get("Most NSCs affected", "Complex V")$estimate, 3)
s1e_most_cv_p   <- fmt_p(s1e_get("Most NSCs affected", "Complex V")$p.value)
s1e_few_ci_p    <- fmt_p(s1e_get("Few NSCs affected",  "Complex I")$p.value)
s1e_few_cv_p    <- fmt_p(s1e_get("Few NSCs affected",  "Complex V")$p.value)
```

In the Most NSCs panel, both knockdowns significantly reduced EdU relative to Control: Complex I (`r s1e_most_ci_est` pp, $p = `r s1e_most_ci_p`$) and Complex V (`r s1e_most_cv_est` pp, $p = `r s1e_most_cv_p`$). Neither contrast was significant in the Few NSCs panel (Complex I $p = `r s1e_few_ci_p`$; Complex V $p = `r s1e_few_cv_p`$), consistent with the non-significant ANOVA for that panel.

---

## Fig S7C: pAMPK levels and ATP depletion

Normalised pAMPK intensity is compared across Control, Complex I,
and Complex V RNAi, one NSC per row, with brain as a random intercept. The
response is log-transformed:

`log(normalized_intensity) ~ condition + (1 | brain_uid)`

Complex I and Complex V RNAi are each compared to Control RNAi, together with a
direct Complex I vs. Complex V comparison, jointly multivariate-$t$ adjusted.
Contrasts are back-transformed and reported as fold-changes relative to Control
RNAi.

```{r}
#| label: pampk-load-data
#| code-summary: "Load Fig S7C pAMPK data and tabulate NSCs and brains per condition"
#| output: true

pampk_df <- read_stat("fig_s07c_pampk.csv") %>%
    filter(!is.na(value)) %>%
    transmute(
        normalized_intensity = value,
        condition = factor(genotype,
                           levels = c("Control RNAi", "Complex I RNAi", "Complex V RNAi")),
        brain_uid = factor(replicate_id)
    )

pampk_df %>%
    group_by(condition) %>%
    summarise(
        n_measurements = n(),
        n_brains       = n_distinct(brain_uid),
        .groups        = "drop"
    ) %>%
    minimal_tbl()
```

### Diagnostics

```{r}
#| label: pampk-lme
#| code-summary: "Fit brain-level linear mixed-effects model"

pampk_model <- lmerTest::lmer(
    log(normalized_intensity) ~ condition + (1 | brain_uid),
    data = pampk_df,
    REML = TRUE
)
summary(pampk_model)
```

::: {.callout-note appearance="simple" collapse="true"}
## Model fit diagnostics with DHARMa

Scaled residuals from 1,000 simulations are tested for uniform distribution (Kolmogorov–Smirnov), dispersion, and an excess of outliers; all three should be non-significant for a well-fitting model.

```{r}
#| label: pampk-dharma
#| code-summary: "Run DHARMa diagnostics"
#| output: true
#| fig-width: 8
#| fig-height: 5

sim_pampk <- DHARMa::simulateResiduals(pampk_model, n = 1000, seed = DHARMA_SEED)
plot(sim_pampk)

invisible(capture.output(
    pampk_dharma_tests <- list(
        uniformity     = DHARMa::testUniformity(sim_pampk, plot = FALSE),
        dispersion     = DHARMa::testDispersion(sim_pampk, plot = FALSE),
        outliers = DHARMa::testOutliers(sim_pampk, plot = FALSE)
    )
))

pampk_dharma_tests %>%
    purrr::map_dfr(
        \(t) tibble(statistic = t$statistic, p.value = t$p.value),
        .id = "test"
    ) %>%
    mutate(across(where(is.numeric), \(x) signif(x, 3)), sig = sig_stars(p.value)) %>%
    minimal_tbl()
```

:::

### Global F-test

Type III ANOVA (Satterthwaite degrees of freedom) on the `condition` term.

```{r}
#| label: pampk-anova
#| code-summary: "Type III ANOVA"
#| output: true

pampk_anova <- anova(pampk_model, type = "III", ddf = "Satterthwaite") %>%
    as.data.frame()

pampk_anova %>%
    tibble::rownames_to_column("term") %>%
    rename(df_num = NumDF, df_den = DenDF, F = `F value`, p.value = `Pr(>F)`) %>%
    mutate(across(where(is.numeric), \(x) signif(x, 3)), sig = sig_stars(p.value)) %>%
    sig_tbl()
```

```{r}
#| label: pampk-anova-stats
#| echo: false
#| output: false
pampk_f    <- signif(pampk_anova["condition", "F value"], 3)
pampk_dfn  <- pampk_anova["condition", "NumDF"]
pampk_dfd  <- signif(pampk_anova["condition", "DenDF"], 3)
pampk_anova_p <- fmt_p(pampk_anova["condition", "Pr(>F)"])
```

The `condition` main effect is significant ($F(`r pampk_dfn`, `r pampk_dfd`) = `r pampk_f`$, $p = `r pampk_anova_p`$); most of the variance is between brains rather than within them.

### All pairwise comparisons (multivariate-$t$ adjusted)

Contrasts are computed on the log scale and back-transformed, so `ratio` is the
multiplicative change in pAMPK intensity between conditions.

```{r}
#| label: pampk-contrasts
#| code-summary: "Pairwise RNAi contrasts, back-transformed to fold-changes (multivariate-$t$ adjusted)"
#| output: true

pampk_em <- emmeans::emmeans(pampk_model, specs = ~ condition)

pampk_contrast_list <- list(
    "Complex I RNAi - Control RNAi"   = c(-1,  1,  0),
    "Complex V RNAi - Control RNAi"   = c(-1,  0,  1),
    "Complex I RNAi - Complex V RNAi" = c( 0,  1, -1)
)

pampk_contrast_results <- emmeans::contrast(
    pampk_em,
    method = pampk_contrast_list,
    adjust = "mvt",
    type   = "response"
) %>%
    summary(infer = c(TRUE, TRUE)) %>%
    as_tibble() %>%
    mutate(across(where(is.numeric), \(x) signif(x, 3)), sig = sig_stars(p.value))

pampk_contrast_results %>%
    sig_tbl()
```

```{r}
#| label: pampk-contrast-stats
#| echo: false
#| output: false
pampk_get <- function(a, b) {
    pampk_contrast_results %>% filter(str_detect(contrast, fixed(a)), str_detect(contrast, fixed(b)))
}
pampk_ci_ctrl_ratio <- pampk_get("Complex I RNAi", "Control")$ratio
pampk_ci_ctrl_p     <- fmt_p(pampk_get("Complex I RNAi", "Control")$p.value)
pampk_cv_ctrl_ratio <- pampk_get("Complex V RNAi", "Control")$ratio
pampk_cv_ctrl_p     <- fmt_p(pampk_get("Complex V RNAi", "Control")$p.value)
pampk_ci_cv_ratio   <- pampk_get("Complex I RNAi", "Complex V")$ratio
pampk_ci_cv_p       <- fmt_p(pampk_get("Complex I RNAi", "Complex V")$p.value)
```

Complex I RNAi reduces pAMPK relative to Control RNAi (ratio = `r pampk_ci_ctrl_ratio`, $p = `r pampk_ci_ctrl_p`$); Complex V RNAi does not differ significantly from Control RNAi (ratio = `r pampk_cv_ctrl_ratio`, $p = `r pampk_cv_ctrl_p`$). Once brain clustering is accounted for, Complex I RNAi and Complex V RNAi are no longer significantly distinguishable from each other (ratio = `r pampk_ci_cv_ratio`, $p = `r pampk_ci_cv_p`$).

---

## Fig S7F: SOD2 rescue experiment

NSC mitotic index (% pH3⁺ among Dpn⁺) is compared in a 3 × 2
factorial of RNAi condition (`Control`, `Complex I`, `Complex V`) × SOD2
expression (`None`, `SOD2`), one brain per row (5–7 per cell). An interaction
linear model is fit to the per-brain values:

`Mitotic_index ~ rnai * sod2`

```{r}
#| label: sod2-load-data
#| code-summary: "Load Fig S7F SOD2-rescue data and tabulate brains per cell"
#| output: true

sod2_df <- read_stat("fig_s07f_sod2_rescue.csv") %>%
    filter(!is.na(value)) %>%
    transmute(
        Mitotic_index = value,
        sod2 = factor(rescue_transgene, levels = c("None", "SOD2")),
        rnai = factor(rnai,             levels = c("Control", "Complex I", "Complex V"))
    )

contrasts(sod2_df$sod2) <- contr.sum(nlevels(sod2_df$sod2))
contrasts(sod2_df$rnai) <- contr.sum(nlevels(sod2_df$rnai))

sod2_df %>%
    count(rnai, sod2, name = "n_brains") %>%
    minimal_tbl()
```

### Diagnostics

```{r}
#| label: sod2-lm-raw
#| code-summary: "Fit interaction linear model (RNAi × SOD2) on per-brain mitotic index"

sod2_lm_raw <- lm(Mitotic_index ~ rnai * sod2, data = sod2_df)
summary(sod2_lm_raw)
```

::: {.callout-note appearance="simple" collapse="true"}
## Model fit diagnostics with DHARMa

Scaled residuals from 1,000 simulations are tested for uniform distribution (Kolmogorov–Smirnov), dispersion, and an excess of outliers; all three should be non-significant for a well-fitting model.

```{r}
#| label: sod2-dharma-raw
#| code-summary: "Run DHARMa diagnostics"
#| output: true
#| fig-width: 8
#| fig-height: 5

sim_sod2_raw <- DHARMa::simulateResiduals(sod2_lm_raw, n = 1000, seed = DHARMA_SEED)
plot(sim_sod2_raw)

invisible(capture.output(
    sod2_dharma_tests <- list(
        uniformity     = DHARMa::testUniformity(sim_sod2_raw, plot = FALSE),
        dispersion     = DHARMa::testDispersion(sim_sod2_raw, plot = FALSE),
        outliers = DHARMa::testOutliers(sim_sod2_raw, plot = FALSE)
    )
))

sod2_dharma_tests %>%
    purrr::map_dfr(
        \(t) tibble(statistic = t$statistic, p.value = t$p.value),
        .id = "test"
    ) %>%
    mutate(across(where(is.numeric), \(x) signif(x, 3)), sig = sig_stars(p.value)) %>%
    minimal_tbl()
```

:::

### Global F-test

Type III ANOVA on `rnai * sod2`: the two main effects and their interaction.

```{r}
#| label: sod2-anova
#| code-summary: "Type III ANOVA (main effects and interaction)"
#| output: true

car::Anova(sod2_lm_raw, type = "III") %>%
    as.data.frame() %>%
    tibble::rownames_to_column("term") %>%
    rename(df = Df, F = `F value`, p.value = `Pr(>F)`) %>%
    mutate(across(where(is.numeric), \(x) signif(x, 3)), sig = sig_stars(p.value)) %>%
    sig_tbl()
```

### SOD2 vs. no SOD2, within each RNAi background (multivariate-$t$ adjusted)

The three within-background comparisons are taken from the full `rnai * sod2`
estimated-marginal-means grid, with a multivariate-$t$ adjustment applied jointly
across all three.

```{r}
#| label: sod2-contrasts
#| code-summary: "SOD2 vs. no SOD2 within each RNAi background (multivariate-$t$ adjusted)"
#| output: true

sod2_em <- emmeans::emmeans(sod2_lm_raw, ~ rnai * sod2)

sod2_contrast_list <- list(
    "SOD2 - None | Control"   = c(-1,  0,  0,  1,  0,  0),
    "SOD2 - None | Complex I" = c( 0, -1,  0,  0,  1,  0),
    "SOD2 - None | Complex V" = c( 0,  0, -1,  0,  0,  1)
)

sod2_contrast_results <- emmeans::contrast(
    sod2_em,
    method = sod2_contrast_list,
    adjust = "mvt"
) %>%
    summary(infer = c(TRUE, TRUE)) %>%
    as_tibble() %>%
    mutate(across(where(is.numeric), \(x) signif(x, 3)), sig = sig_stars(p.value))

sod2_contrast_results %>%
    sig_tbl()
```

---

## Fig S9A: Glial OxPhos knockdown and NSC background

EdU-positive progeny per NSC clone is compared across six genotypes: the two reference controls (`mCherry x Luciferase`, wild type;
`ND42 x Luciferase`, CI-dysfunction baseline) and the four combinations of NSC
background (`mCherry` = wild type, `ND42` = CI-dysfunction) × glial knockdown
(`ND42` or `ATPsynO`), one clone per row, with brain as the biological
replicate.

The data are fit with a factorial linear mixed-effects model:

`EdU ~ nsc_background * glial_gene + (1 | brain_uid)`

The `nsc_background × glial_gene` interaction tests whether the glial-knockdown
effect depends on NSC background. Two simple-effect comparisons:
`ND42 x ND42` vs. `mCherry x ND42` and `ND42 x ATPsynO` vs. `mCherry x ATPsynO`
test whether NSC background changes the effect of each glial knockdown, jointly
multivariate-$t$ adjusted.

```{r}
#| label: s8e-ox-load-data
#| code-summary: "Load Fig S9A glial-OxPhos data and tabulate clones and brains per genotype"
#| output: true

s8e_ox_target <- c(
    "mCherry x Luciferase", "ND42 x Luciferase",
    "mCherry x ND42", "ND42 x ND42",
    "mCherry x ATPsynO", "ND42 x ATPsynO"
)

s8e_ox_df <- read_stat("edu_glial_screen.csv") %>%
    filter(genotype %in% s8e_ox_target, !is.na(value)) %>%
    transmute(
        EdU      = value,
        Genotype = genotype,
        nsc_background = case_when(
            genotype %in% c("mCherry x Luciferase", "mCherry x ND42", "mCherry x ATPsynO") ~ "WT",
            genotype %in% c("ND42 x Luciferase", "ND42 x ND42", "ND42 x ATPsynO") ~ "CI-dysfunction"
        ),
        glial_gene = case_when(
            genotype %in% c("mCherry x Luciferase", "ND42 x Luciferase") ~ "control",
            genotype %in% c("mCherry x ND42", "ND42 x ND42") ~ "ND42",
            genotype %in% c("mCherry x ATPsynO", "ND42 x ATPsynO") ~ "ATPsynO"
        ),
        nsc_background = factor(nsc_background, levels = c("WT", "CI-dysfunction")),
        glial_gene     = factor(glial_gene, levels = c("control", "ND42", "ATPsynO")),
        brain_uid      = factor(replicate_id)
    )

stopifnot(!anyNA(s8e_ox_df$nsc_background), !anyNA(s8e_ox_df$glial_gene))

s8e_ox_df %>%
    group_by(nsc_background, glial_gene) %>%
    summarise(
        n_clones = n(),
        n_brains = n_distinct(brain_uid),
        .groups  = "drop"
    ) %>%
    minimal_tbl()
```

### Diagnostics

```{r}
#| label: s8e-ox-lme
#| code-summary: "Fit clone-level linear mixed-effects model"

contrasts(s8e_ox_df$nsc_background) <- contr.sum(2)
contrasts(s8e_ox_df$glial_gene) <- contr.sum(3)

s8e_ox_lme <- lmerTest::lmer(
    EdU ~ nsc_background * glial_gene + (1 | brain_uid),
    data = s8e_ox_df,
    REML = TRUE
)
summary(s8e_ox_lme)
```

::: {.callout-note appearance="simple" collapse="true"}
## Model fit diagnostics with DHARMa

Scaled residuals from 1,000 simulations are tested for uniform distribution (Kolmogorov–Smirnov), dispersion, and an excess of outliers; all three should be non-significant for a well-fitting model.

```{r}
#| label: s8e-ox-dharma
#| code-summary: "Run DHARMa diagnostics"
#| output: true
#| fig-width: 8
#| fig-height: 5

sim_s8e_ox <- DHARMa::simulateResiduals(s8e_ox_lme, n = 1000, seed = DHARMA_SEED)
plot(sim_s8e_ox)

invisible(capture.output(
    s8e_ox_dharma_tests <- list(
        uniformity     = DHARMa::testUniformity(sim_s8e_ox, plot = FALSE),
        dispersion     = DHARMa::testDispersion(sim_s8e_ox, plot = FALSE),
        outliers       = DHARMa::testOutliers(sim_s8e_ox, plot = FALSE)
    )
))

s8e_ox_dharma_tests %>%
    purrr::map_dfr(
        \(t) tibble(statistic = t$statistic, p.value = t$p.value),
        .id = "test"
    ) %>%
    mutate(across(where(is.numeric), \(x) signif(x, 3)), sig = sig_stars(p.value)) %>%
    minimal_tbl()
```

:::

### Global F-test

Type III ANOVA (Satterthwaite degrees of freedom) on `nsc_background * glial_gene`: the two main effects and their interaction.

```{r}
#| label: s8e-ox-anova
#| code-summary: "Type III ANOVA (main effects and interaction)"
#| output: true

s8e_ox_anova <- anova(s8e_ox_lme, type = "III", ddf = "Satterthwaite") %>%
    as.data.frame()

s8e_ox_anova %>%
    tibble::rownames_to_column("term") %>%
    rename(df_num = NumDF, df_den = DenDF, F = `F value`, p.value = `Pr(>F)`) %>%
    mutate(across(where(is.numeric), \(x) signif(x, 3)), sig = sig_stars(p.value)) %>%
    sig_tbl()
```

```{r}
#| label: s8e-ox-anova-stats
#| echo: false
#| output: false
s8e_ox_nsc_f   <- signif(s8e_ox_anova["nsc_background", "F value"], 3)
s8e_ox_nsc_dfd <- signif(s8e_ox_anova["nsc_background", "DenDF"], 3)
s8e_ox_nsc_p   <- fmt_p(s8e_ox_anova["nsc_background", "Pr(>F)"])
s8e_ox_glia_f   <- signif(s8e_ox_anova["glial_gene", "F value"], 3)
s8e_ox_glia_dfd <- signif(s8e_ox_anova["glial_gene", "DenDF"], 3)
s8e_ox_glia_p   <- fmt_p(s8e_ox_anova["glial_gene", "Pr(>F)"])
s8e_ox_int_f   <- signif(s8e_ox_anova["nsc_background:glial_gene", "F value"], 3)
s8e_ox_int_dfd <- signif(s8e_ox_anova["nsc_background:glial_gene", "DenDF"], 3)
s8e_ox_int_p   <- fmt_p(s8e_ox_anova["nsc_background:glial_gene", "Pr(>F)"])
```

Both main effects are significant: `nsc_background` ($F(1, `r s8e_ox_nsc_dfd`) = `r s8e_ox_nsc_f`$, $p = `r s8e_ox_nsc_p`$) and `glial_gene` ($F(2, `r s8e_ox_glia_dfd`) = `r s8e_ox_glia_f`$, $p = `r s8e_ox_glia_p`$), confirming that CI-dysfunction NSCs proliferate less overall and that glial OxPhos knockdown reduces proliferation overall. The interaction is not significant ($F(2, `r s8e_ox_int_dfd`) = `r s8e_ox_int_f`$, $p = `r s8e_ox_int_p`$).

### NSC background within each glial knockdown (multivariate-$t$ adjusted)

These test whether NSC background affects proliferation within each
glial-knockdown condition, taken from the full `nsc_background * glial_gene`
estimated-marginal-means grid and jointly multivariate-$t$ adjusted.

```{r}
#| label: s8e-ox-contrasts
#| code-summary: "NSC-background simple effects within each glial knockdown (multivariate-$t$ adjusted)"
#| output: true

s8e_ox_em <- emmeans::emmeans(s8e_ox_lme, ~ nsc_background * glial_gene)

s8e_ox_contrast_list <- list(
    "ND42 x ND42 - mCherry x ND42"       = c(0, 0, -1, 1, 0, 0),
    "ND42 x ATPsynO - mCherry x ATPsynO" = c(0, 0, 0, 0, -1, 1)
)

s8e_ox_contrast_results <- emmeans::contrast(
    s8e_ox_em,
    method = s8e_ox_contrast_list,
    adjust = "mvt"
) %>%
    summary(infer = c(TRUE, TRUE)) %>%
    as_tibble() %>%
    mutate(across(where(is.numeric), \(x) signif(x, 3)), sig = sig_stars(p.value))

s8e_ox_contrast_results %>%
    sig_tbl()
```

```{r}
#| label: s8e-ox-contrast-stats
#| echo: false
#| output: false
s8e_ox_get <- function(gene) s8e_ox_contrast_results %>% filter(str_detect(contrast, fixed(gene)))
s8e_ox_nd42_est <- s8e_ox_get("x ND42 -")$estimate
s8e_ox_nd42_p   <- fmt_p(s8e_ox_get("x ND42 -")$p.value)
s8e_ox_atp_est  <- s8e_ox_get("ATPsynO -")$estimate
s8e_ox_atp_p    <- fmt_p(s8e_ox_get("ATPsynO -")$p.value)
```

Both comparisons trend in the expected direction, CI-dysfunction NSCs proliferate less than wild-type NSCs when the glia also has `ND42` (`r s8e_ox_nd42_est` EdU pp, $p = `r s8e_ox_nd42_p`$) or `ATPsynO` (`r s8e_ox_atp_est` EdU pp, $p = `r s8e_ox_atp_p`$) knocked down, but neither reaches significance.

### Glial knockdown vs. pooled controls, within each NSC background (multivariate-$t$ adjusted)

These compare each glial-knockdown genotype against the pooled `Luciferase`
control *within* the matching NSC background, testing whether glial knockdown
affects wild-type and CI-dysfunction NSCs differently. All four comparisons are
taken from the `nsc_background * glial_gene` grid and jointly multivariate-$t$
adjusted.

```{r}
#| label: s8e-ox-contrasts-vs-control
#| code-summary: "Glial knockdown vs. pooled control within each NSC background (multivariate-$t$ adjusted)"
#| output: true

s8e_ox_vs_control_list <- list(
    "mCherry x ND42 - mCherry x Luciferase"    = c(-1,  0,  1,  0,  0,  0),
    "mCherry x ATPsynO - mCherry x Luciferase" = c(-1,  0,  0,  0,  1,  0),
    "ND42 x ND42 - ND42 x Luciferase"          = c( 0, -1,  0,  1,  0,  0),
    "ND42 x ATPsynO - ND42 x Luciferase"       = c( 0, -1,  0,  0,  0,  1)
)

s8e_ox_vs_control_results <- emmeans::contrast(
    s8e_ox_em,
    method = s8e_ox_vs_control_list,
    adjust = "mvt"
) %>%
    summary(infer = c(TRUE, TRUE)) %>%
    as_tibble() %>%
    mutate(across(where(is.numeric), \(x) signif(x, 3)), sig = sig_stars(p.value))

s8e_ox_vs_control_results %>%
    sig_tbl()
```

```{r}
#| label: s8e-ox-contrast-vs-control-stats
#| echo: false
#| output: false
s8e_ox_get_vc <- function(pattern) s8e_ox_vs_control_results %>% filter(str_detect(contrast, pattern))
s8e_ox_wt_nd42_est  <- s8e_ox_get_vc(fixed("mCherry x ND42 -"))$estimate
s8e_ox_wt_nd42_p    <- fmt_p(s8e_ox_get_vc(fixed("mCherry x ND42 -"))$p.value)
s8e_ox_wt_atp_est   <- s8e_ox_get_vc(fixed("mCherry x ATPsynO -"))$estimate
s8e_ox_wt_atp_p     <- fmt_p(s8e_ox_get_vc(fixed("mCherry x ATPsynO -"))$p.value)
s8e_ox_ci_nd42_est  <- s8e_ox_get_vc(fixed("ND42 x ND42 -"))$estimate
s8e_ox_ci_nd42_p    <- fmt_p(s8e_ox_get_vc(fixed("ND42 x ND42 -"))$p.value)
s8e_ox_ci_atp_est   <- s8e_ox_get_vc(fixed("ND42 x ATPsynO -"))$estimate
s8e_ox_ci_atp_p     <- fmt_p(s8e_ox_get_vc(fixed("ND42 x ATPsynO -"))$p.value)
```

Within the wild-type NSC background, neither glial knockdown reaches
significance relative to `Luciferase` control (`ND42`: `r s8e_ox_wt_nd42_est` EdU pp, $p = `r s8e_ox_wt_nd42_p`$; `ATPsynO`: `r s8e_ox_wt_atp_est` EdU pp, $p = `r s8e_ox_wt_atp_p`$), though the `ND42` comparison trends toward a reduction. Within the CI-dysfunction NSC background, both glial knockdowns significantly reduce proliferation relative to `Luciferase` control (`ND42`: `r s8e_ox_ci_nd42_est` EdU pp, $p = `r s8e_ox_ci_nd42_p`$; `ATPsynO`: `r s8e_ox_ci_atp_est` EdU percentage-points, $p = `r s8e_ox_ci_atp_p`$).

---

## Fig S9I: Glial LbNox and CI dysfunction rescue

EdU-positive progeny per NSC clone is compared across three genotypes differing
in glial LbNox expression and NSC metabolic state, one clone per row, with brain
as the biological replicate. VNCs from different imaging sessions or genotypes
are counted as distinct brains. The three genotypes are:

  - `mCherry x Luciferase` ($n = 78$, 16 brains): shared WT control pool.
  - `mCherry x LbNox` ($n = 59$, 11 brains): glial LbNox in a WT NSC background,
    testing whether glial LbNox alone affects proliferation.
  - `ND42 x LbNox` ($n = 34$, 7 brains): glial LbNox with CI-deficient NSCs,
    testing whether NSC metabolic state modifies the glial LbNox effect.

All three conditions are fit in one clone-level linear mixed-effects model with
brain as a random intercept, `EdU ~ condition + (1 | brain_uid)`.

Three comparisons, specified a priori from the experimental design, are tested
together with a joint multivariate-$t$ adjustment:

1. `mCherry x LbNox` vs `mCherry x Luciferase`: does glial LbNox reduce WT NSC
proliferation?
2. `ND42 x LbNox` vs `mCherry x LbNox`: is the LbNox effect reversed in
CI-deficient NSCs?
3. `ND42 x LbNox` vs `mCherry x Luciferase`: is proliferation in CI-deficient
NSCs with glial LbNox restored to the WT level?

```{r}
#| label: s8j-load-data
#| code-summary: "Load Fig S9I glial-LbNox data and tabulate clones and brains per condition"
#| output: true

LUC_WT_S9I   <- "mCherry x Luciferase"
LBNOX_WT_S9I <- "mCherry x LbNox"
LBNOX_CI_S9I <- "ND42 x LbNox"

s8j_conditions <- c(LUC_WT_S9I, LBNOX_WT_S9I, LBNOX_CI_S9I)

s8j_df <- read_stat("edu_glial_screen.csv") %>%
    filter(!is.na(value), genotype %in% s8j_conditions) %>%
    transmute(
        EdU       = value,
        date      = collection_date,
        brain_uid = replicate_id,
        condition = factor(genotype, levels = s8j_conditions)
    )

s8j_df %>%
    group_by(condition) %>%
    summarise(
        n_clones = n(),
        n_brains = n_distinct(brain_uid),
        n_dates  = n_distinct(date),
        .groups  = "drop"
    ) %>%
    minimal_tbl()
```

### Diagnostics

```{r}
#| label: s8j-lme-fit
#| code-summary: "Fit linear mixed-effects model"

s8j_model <- lmerTest::lmer(
    EdU ~ condition + (1 | brain_uid),
    data    = s8j_df,
    REML    = TRUE,
    control = lme4::lmerControl(optimizer = "bobyqa")
)

summary(s8j_model)
```

::: {.callout-note appearance="simple" collapse="true"}
## Model fit diagnostics with DHARMa

Scaled residuals from 1,000 simulations are tested for uniform distribution (Kolmogorov–Smirnov), dispersion, and an excess of outliers; all three should be non-significant for a well-fitting model.

```{r}
#| label: s8j-dharma
#| code-summary: "Run DHARMa diagnostics"
#| output: true
#| fig-width: 8
#| fig-height: 5

sim_s8j <- DHARMa::simulateResiduals(s8j_model, n = 1000, seed = DHARMA_SEED)
plot(sim_s8j)

invisible(capture.output(
    s8j_dharma_tests <- list(
        uniformity    = DHARMa::testUniformity(sim_s8j, plot = FALSE),
        dispersion    = DHARMa::testDispersion(sim_s8j, plot = FALSE),
        outliers = DHARMa::testOutliers(sim_s8j, plot = FALSE)
    )
))

s8j_dharma_tests %>%
    purrr::map_dfr(
        \(t) tibble(statistic = t$statistic, p.value = t$p.value),
        .id = "test"
    ) %>%
    mutate(across(where(is.numeric), \(x) signif(x, 3)), sig = sig_stars(p.value)) %>%
    minimal_tbl()
```

:::

### Global F-test

Type III ANOVA (Satterthwaite degrees of freedom) on the `condition` term.

```{r}
#| label: s8j-anova
#| code-summary: "Type III ANOVA"
#| output: true

s8j_anova_res <- anova(s8j_model, type = "III", ddf = "Satterthwaite") %>%
    as.data.frame()

s8j_anova_res %>%
    tibble::rownames_to_column("term") %>%
    rename(df_num = NumDF, df_den = DenDF, F = `F value`, p.value = `Pr(>F)`) %>%
    mutate(across(where(is.numeric), \(x) signif(x, 3)), sig = sig_stars(p.value)) %>%
    sig_tbl()
```

```{r}
#| label: s8j-anova-stats
#| echo: false
#| output: false
s8j_f   <- signif(s8j_anova_res["condition", "F value"], 3)
s8j_dfn <- s8j_anova_res["condition", "NumDF"]
s8j_dfd <- signif(s8j_anova_res["condition", "DenDF"], 3)
s8j_p   <- fmt_p(s8j_anova_res["condition", "Pr(>F)"])
```

The `condition` term was not significant ($F(`r s8j_dfn`, `r s8j_dfd`) = `r s8j_f`$, $p = `r s8j_p`$).

### Glial LbNox effect and proliferation restoration in CI-deficient NSCs

All estimates are in EdU percentage points.

```{r}
#| label: s8j-emmeans
#| code-summary: "Estimated marginal means per condition"

s8j_em <- emmeans::emmeans(s8j_model, specs = ~ condition)
summary(s8j_em)
```

```{r}
#| label: s8j-contrasts
#| code-summary: "The three LbNox contrasts (multivariate-$t$ adjusted)"
#| output: true

s8j_contrast_list <- list(
    "mCherry x LbNox - mCherry x Luciferase" = c(-1,  1,  0),
    "ND42 x LbNox - mCherry x LbNox"          = c( 0, -1,  1),
    "ND42 x LbNox - mCherry x Luciferase"     = c(-1,  0,  1)
)

s8j_contrasts_summary <- emmeans::contrast(s8j_em, method = s8j_contrast_list, adjust = "mvt") %>%
    summary(infer = c(TRUE, TRUE)) %>%
    as_tibble()

s8j_contrasts_summary %>%
    mutate(across(where(is.numeric), \(x) signif(x, 3)), sig = sig_stars(p.value)) %>%
    sig_tbl()
```

```{r}
#| label: s8j-contrast-stats
#| echo: false
#| output: false
s8j_key     <- s8j_contrasts_summary %>%
    filter(str_detect(contrast, "ND42 x LbNox - mCherry x LbNox"))
s8j_key_est <- signif(s8j_key$estimate[1], 3)
s8j_key_t   <- signif(s8j_key$t.ratio[1], 3)
s8j_key_df  <- signif(s8j_key$df[1], 3)
s8j_key_p   <- fmt_p(s8j_key$p.value[1])
s8j_all_ns  <- all(s8j_contrasts_summary$p.value >= 0.05)
```

All three contrasts are non-significant. The key contrast (`ND42 x LbNox` − `mCherry x LbNox`, `r s8j_key_est` EdU pp, $t(`r s8j_key_df`) = `r s8j_key_t`$, $p = `r s8j_key_p`$) is in the expected direction but does not survive multivariate-$t$ correction.

---

```{r}
#| label: write-all-results
#| echo: false
#| output: false

# Machine-readable copies of every result table in this report, written to
# results/statistical_analysis/ (git-ignored, regenerated on each render).
write_result(fig1b_contrast_results,            "fig_01b_results.csv")
write_result(fig1g_anova,                       "fig_01g_anova.csv")
write_result(fig1p_contrast_results,            "fig_01p_results.csv")
write_result(fig1p_anova,                       "fig_01p_anova.csv")
write_result(ucp_contrasts_summary,             "fig_03e_within_condition_results.csv")
write_result(ucp_direct_interactions_summary,   "fig_03e_diff_in_diff_results.csv")
write_result(lbnox_contrasts_summary,           "fig_03g_within_condition_results.csv")
write_result(lbnox_direct_interactions_summary, "fig_03g_diff_in_diff_results.csv")
write_result(inx2_contrast_results,             "fig_04c_results.csv")
write_result(combined_results,                  "combined_screen_results.csv")
write_result(phenotype_contrast_summary,        "combined_phenotype_contrast.csv")
write_result(eaat1_left_contrast_summary,       "fig_04k_left_results.csv")
write_result(eaat1_right_contrast_summary,      "fig_04k_right_results.csv")
write_result(ldh_alone_summary,                 "fig_05e_ldh_alone.csv")
write_result(ldh_worsens_summary,               "fig_05e_ldh_worsens.csv")
write_result(ldh_rescue_vs_baseline_summary,    "fig_05e_rescue_vs_baseline.csv")
write_result(ldh_rescue_vs_control_summary,     "fig_05e_rescue_vs_control.csv")
write_result(s1e_contrast_results,              "fig_s01i_j_results.csv")
write_result(pampk_contrast_results,            "fig_s07c_results.csv")
write_result(sod2_contrast_results,             "fig_s07f_results.csv")
write_result(s8e_ox_contrast_results,           "fig_s09a_results.csv")
write_result(s8e_ox_vs_control_results,         "fig_s09a_vs_control_results.csv")
write_result(s8j_contrasts_summary,             "fig_s09i_results.csv")
```

## Software and Statistical Analysis

All analyses were performed in R using linear and generalized least-squares
models (`stats::lm`, `nlme::gls`), linear mixed-effects models
(`lme4`, `lmerTest`), estimated marginal means and multiplicity-corrected
contrasts (`emmeans`), Type III ANOVA (`car`, `lmerTest`), and simulation-based
residual diagnostics (`DHARMa`). Data handling used the `tidyverse`; tables and
figures used `kableExtra` and `reactable`. The per-analysis data tables are in
`statistical_analysis/data/`, one row per clone, brain, or cell.

Reported p-values are two-sided. Small, pre-specified contrast sets use a
multivariate-$t$ adjustment; the large glial RNAi screen uses a Benjamini-Hochberg
FDR correction. A fixed random seed is set for the model diagnostics and for the
multivariate-$t$ computation, so all reported values are reproducible. Complete
package versions are given in the Session Information below.

::: {.callout-note appearance="simple" collapse="true"}
## Session Information

```{r}
#| label: session-info
#| output: true
#| echo: false

sessionInfo()
```

:::
